---
title: "Understanding Interactions in Regression Models: Part 3 - Generalized Linear Models"
description: "Second in a series on interactions: exploring numerical × numerical interactions, marginal effects, conditional effects, and distributional regression in R"
author:
  - name: Allen Bush-Beaupré
    url: https://abushbeaupre.github.io/
    orcid: 0000-0001-6989-9278
date: 07-27-2026
categories: [tutorial, interactions, marginaleffects, glmm, statistics, R, series]
draft: true
editor: source
format:
  html:
    fig-width: 10
    fig-height: 7
    out-width: "100%"
---
# Introduction

In this third part of my tutorial series on statistical interactions, we are going to push further on our understanding and reporting of interactions. Since we are going beyond the realm of linear models by exploring generalized linear models (GLMs) and their variants, I am assuming here that the reader has a basic understanding of how GLMs work especially concerning their use of link functions.

I typically write up these tutorials based on bits and pieces of information I have gathered here and there but I will (in part) address topics covered by Spake et al. (2023) https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12939 while giving examples of tools that can be used to report interaction effects on the additive and multiplicative scale.

The case studies in this tutorial will focus on species occurence and abundance. We will first look at habitat suitability for a given species to occur in a given location. Second, we will explore factors that affect their abundance when they are present. Third, we will incorporate both these processes in a single model and learn how to extract as much information as possible from these types of analyses.

## Setup

First, we load the relevant packages and set up our ggplot2 theme for the rest of the document.

```{r}
#| output: false
#| warning: false
library(tidyverse)       # Data manipulation and plotting
library(tidylog)         # Print a log of data manipulations done with tidyverse
library(glmmTMB)         # Generalized linear (mixed) models
library(marginaleffects) # Compute marginal/conditional effects
library(patchwork)       # Combine multiple plots
library(distributional)  # Simulate distributions
library(ggdist)          # Plot nice distributions
library(GGally)

# Set ggplot2 theme
black_theme <- theme(
  axis.line = element_line(linewidth = 2, lineend = "round", color = "white"),
  panel.grid = element_blank(),
  panel.background = element_rect(fill = "black", color = NA),
  axis.ticks = element_blank(),
  axis.text = element_text(size = 21, face = "bold", color = "white"),
  axis.title = element_text(size = 21, face = "bold", color = "white"),
  plot.title = element_text(size = 25, face = "bold", color = "white"),
  plot.background = element_rect(fill = "black", color = NA),
  legend.background = element_rect(fill = "black", color = NA),
  legend.text = element_text(size = 21, color = "white"),
  legend.title = element_text(size = 21, face = "bold", color = "white"),
  legend.position = "bottom"
)

theme_set(black_theme)

# Set global default for geom_point shape
update_geom_defaults("point", list(shape = 1, size = 3, alpha = 0.8))

# silence marginaleffects warnings about glmmTMB
options(marginaleffects_safe = FALSE)

set.seed(333)


```

# Case Study 1: Species occurence
In this first case study, we are interested in species occurence in different habitats. Specifically, we want evaluate the the interacting effects of habitat composition and environmental variables on the probability of occurence in a given patch. 


https://ben18785.shinyapps.io/distribution-zoo/ 

```{r}
n_plots <- 200

tree_cover <- dist_mixture(dist_beta(4, 0.8), dist_beta(0.8, 4), weights = c(0.5, 0.5))

avg_yearly_rain <- dist_gamma(shape = 3, rate = 0.7)

avg_yearly_temp <- dist_normal(mu = 20, sd = 10)

occurence_intercept <- plogis(-3)

occurence_tree_effect <- 0.45

occurence_rain_effect <- 0.01

occurence_temp_effect <- 0.01

tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |> 
  unnest(everything()) |>
  mutate(occurence = rbinom(n_plots, size = 1, 
                     prob = occurence_intercept +
                            occurence_tree_effect * tree_cov +
                            occurence_rain_effect * rain +
                            occurence_temp_effect * temp
                            ))-> occ_data

ggpairs(
  occ_data,
  diag  = list(continuous = wrap("densityDiag", fill = "white", alpha = 0.6, color = "white")),
  lower = list(continuous = wrap("points", color = "white", alpha = 0.6, size = 2, shape = 1)),
  upper = list(continuous = wrap("cor", color = "white", size = 5))
) +
  theme(
    strip.text = element_text(color = "black", face = "bold", size = 12),
    axis.text = element_text(color = "white", face = "bold", size = 12)
  )
```

Some people might want to dichotomize tree cover
```{r}
scale_x <- function(x){
  ( x - mean(x) ) / sd(x)
}


occ_data |>
  mutate(tree_cat = as.factor(ifelse(tree_cov > 0.5, "high", "low")),
         scl_tree = scale_x(tree_cov),
         scl_rain = scale_x(rain),
         scl_temp = scale_x(temp)) -> occ_data
```

Implicit interactions

```{r}
mod_1 <- glmmTMB(occurence ~ tree_cov + rain + temp, family = binomial(link = "logit"), data = occ_data)

summary(mod_1)

slopes(mod_1, variables = "tree_cov", newdata = datagrid(temp = mean(occ_data$temp), rain = mean(occ_data$rain)))
```

Notice something strange: we simulated the effect of tree cover on occurrence to be 0.45, but `summary(mod_1)` reports a slope of about 1.87 for `tree_cov`. Did the model get it wrong? No — it's reporting the effect on a different scale.

When we fit a binomial GLM with a logit link, the model isn't drawing a straight line through probabilities. Instead, it draws a straight line through the *log-odds* of occurrence, and then bends that line into an S-shape (between 0 and 1) to get back to a probability. The number 1.87 is the slope on that log-odds scale. It's a real, meaningful number, but it's not directly comparable to our simulated 0.45, which was the slope on the probability scale.

Here's the key idea: because the logit link bends the line, how much a one-unit increase in tree cover changes the *probability* of occurrence depends on where you are on the curve. Near the middle of the S-shape (probability around 50%), a small push on the log-odds scale translates into a fairly large push on the probability scale. Near the top or bottom of the S-shape (probability near 0% or 100%), that same push on the log-odds scale barely moves the probability at all, because the curve has flattened out.

This is exactly why `slopes(mod_1, variables = "tree", newdata = datagrid(temp = mean(occ_data$temp), rain = mean(occ_data$rain)))` gives us back something close to 0.45. Rather than reporting the abstract log-odds slope, `slopes()` asks: "at these specific values of temp and rain, how much does the predicted *probability* change when tree cover increases a little?" That's a probability-scale answer, so it's directly comparable to the number we used to simulate the data.

There's a subtle but important consequence here: because the amount of "bending" depends on where you are on the S-curve, the probability-scale effect of tree cover also depends on the values of rain and temp — even though we never told the model to include an interaction between them! This is what we mean by an *implicit interaction*. The logit link function automatically makes every predictor's effect (on the probability scale) depend on the values of all the other predictors, simply because of the shape of the S-curve. We'll see this play out visually below, where the estimated slope for tree cover changes depending on the combination of rain and temp we plug in.

Spake et al. (2023, *Biological Reviews*) make a related point in the ecology literature: when we ask "does an interaction exist?", the answer can depend on which *measurement scale* we ask the question on — additive (e.g., probability, or log-odds treated as a straight difference) versus multiplicative (e.g., ratios, or a logarithmic scale). Getting the scale question wrong is one of the ways they describe interactions being mis-detected or mis-sized (what they call a "Type-D" error). Our tree-cover example is a small, concrete illustration of exactly this idea: there is no interaction term of any kind in `mod_1`'s formula, yet whether or not you can *see* an interaction between tree cover and the other predictors depends entirely on whether you're looking at the log-odds scale (no interaction, by construction) or the probability scale (an interaction appears, because of the S-curve).

```{r}
#| echo: false
#| fig-height: 5
tibble(log_odds = seq(-6, 6, length.out = 400)) |>
  mutate(probability = plogis(log_odds)) -> logistic_curve

tibble(
  log_odds = c(0, -4.2, 4.2),
  probability = plogis(log_odds),
  label = c("steep middle:\na small push here\nmoves probability a lot",
            "flat end:\na push here barely\nmoves probability",
            "flat end:\na push here barely\nmoves probability")
) -> annot

ggplot(logistic_curve, aes(x = log_odds, y = probability)) +
  geom_line(color = "white", linewidth = 1.2) +
  geom_point(data = annot, color = "white", size = 3) +
  geom_text(
    data = annot,
    aes(label = label),
    color = "white",
    size = 4.5,
    fontface = "bold",
    lineheight = 0.9,
    vjust = c(-0.4, -0.6, 1.6),
    hjust = c(0.5, 0.5, 0.5)
  ) +
  scale_y_continuous(limits = c(-0.1, 1.05)) +
  labs(
    x = "log-odds scale (what the model's coefficient describes)",
    y = "probability scale (what we actually care about)",
    title = "The logit link bends a straight line into an S-shape"
  )
```


```{r}
slopes(mod_1, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) -> t

ggplot(t) + geom_pointrange(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = 1), color = "white") +
  geom_vline(xintercept = occurence_tree_effect, color = "white", alpha = .8) +
  facet_grid(rain ~ temp)
```


```{r}
predictions(mod_1, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) -> pred_tree1

ggplot(pred_tree1) +
  geom_ribbon(aes(x = tree_cov, ymin = conf.low, ymax = conf.high), fill = "white", alpha = 0.2) +
  geom_line(aes(x = tree_cov, y = estimate), color = "white") +
  facet_grid(rain ~ temp)
```

# Case Study 2: Species abundance

Now let's look at the same three predictors, but for a different question: given that the species is present in a patch, how many individuals do we expect to find there? Counts like this (0, 1, 2, 3, ...) are commonly modeled with a Poisson distribution, using a **log link** instead of the logit link we used for occurrence.

```{r}
abundance_intercept <- exp(1)

abundance_tree_effect <- 4.5

abundance_rain_effect <- 0.1

abundance_temp_effect <- 0.1

tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |> 
  unnest(everything()) |>
  mutate(abundance = rpois(n_plots, lambda = 
                     abundance_intercept +
                     abundance_tree_effect * tree_cov +
                     abundance_rain_effect * rain +
                     abundance_temp_effect * temp
                     ))-> abund_data

ggpairs(
  abund_data,
  diag  = list(continuous = wrap("densityDiag", fill = "white", alpha = 0.6, color = "white")),
  lower = list(continuous = wrap("points", color = "white", alpha = 0.6, size = 2, shape = 1)),
  upper = list(continuous = wrap("cor", color = "white", size = 5))
) +
  theme(
    strip.text = element_text(color = "black", face = "bold", size = 12),
    axis.text = element_text(color = "white", face = "bold", size = 12)
  )
```

Just like before, we simulated the effect of tree cover on abundance to be additive and equal to 4.5 (expected count-scale slope), but we're about to fit a model that assumes an *additive* relationship on the log scale.

```{r}
mod_2 <- glmmTMB(abundance ~ tree_cov + rain + temp, family = poisson(link = "log"), data = abund_data)

summary(mod_2)

slopes(mod_2, variables = "tree_cov", newdata = datagrid(temp = mean(abund_data$temp), rain = mean(abund_data$rain)))
```

The same story repeats itself. `summary(mod_2)` reports a `tree_cov` coefficient that is not 4.5 — because a Poisson GLM with a log link draws a straight line through the *log* of the expected count, not through the expected count itself. To turn a log-scale prediction back into an actual expected count, we exponentiate it, and that exponentiation is what bends a straight line into a curve that always stays positive (counts can never be negative).

Because of that curve, a fixed log-scale coefficient translates into a *different* count-scale slope depending on where you evaluate it — exactly the same phenomenon we saw with the logistic curve. This time, the local rate of change works out to $$\lambda \cdot \beta$$ (the expected count itself, times the log-scale coefficient), instead of $$p(1-p) \cdot \beta$$. When we plug in the average `rain` and `temp` values via `slopes()`, we get back a count-scale slope for `tree_cov` close to our simulated 4.5.

And once again, this produces an *implicit interaction*: the count-scale effect of tree cover depends on the values of rain and temp (through their influence on the expected count $$\lambda$$), even though the model formula never specified an interaction term. Any log-link GLM will behave this way — additive on the log scale automatically becomes multiplicative (and interaction-like) on the response scale.

This is the same additive-versus-multiplicative distinction that Spake et al. (2023, *Biological Reviews*) highlight as a routinely overlooked property of statistical interactions: whether a response is being analyzed on an additive scale or a multiplicative one (such as a ratio or logarithmic scale). A log-link Poisson model is additive on the log scale by construction — that's the whole point of the link function — but that same additivity becomes multiplicative on the count scale, since exponentiating a sum turns it into a product ($$e^{a+b} = e^a \times e^b$$). Two predictors that don't interact at all on the log scale will therefore appear to interact on the count scale, for the same reason the abstract flags as a source of "Type-D" errors: judging whether or how strongly an interaction exists depends on which scale you're looking at.

```{r}
slopes(mod_2, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) -> t2

ggplot(t2) + geom_pointrange(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = 1), color = "white") +
  geom_vline(xintercept = abundance_tree_effect, color = "white", alpha = .8) +
  facet_grid(rain ~ temp)
```

```{r}
predictions(mod_2, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) -> pred_tree2

ggplot(pred_tree2) +
  geom_ribbon(aes(x = tree_cov, ymin = conf.low, ymax = conf.high), fill = "white", alpha = 0.2) +
  geom_line(aes(x = tree_cov, y = estimate), color = "white") +
  facet_grid(rain ~ temp)
```





