---
title: "Understanding Interactions in Regression Models: Part 2 - Numerical × Numerical"
description: "First in a series on interactions: exploring categorical × numerical interactions, marginal effects, Simpson's Paradox, and sampling bias using the marginaleffects package in R"
author:
  - name: Allen Bush-Beaupré
    url: https://abushbeaupre.github.io/
    orcid: 0000-0001-6989-9278
date: 02-25-2026
categories: [tutorial, interactions, marginaleffects, glmm, statistics, R, series]
draft: false
editor: source
format:
  html:
    fig-width: 10
    fig-height: 7
    out-width: "100%"
---

# Introduction

This is a follow-up on my previous post about categorical x numerical interactions in regression models. We will now look at numerical x numerical interactions, how to interpret and visualize them, and add a twist to the typical ways they are included or excluded from models. We will stay with Gaussian-distributed response variables for now and add other distributions in the next tutorial.

## Setup

First, we load the relevant packages and set up our ggplot2 theme for the rest of the document.

```{r}
#| output: false
#| warning: false
library(tidyverse)       # Data manipulation and plotting
library(tidylog)         # Print a log of data manipulations done with tidyverse
library(glmmTMB)         # Generalized linear (mixed) models
library(marginaleffects) # Compute marginal/condition effects
library(patchwork)       # Combine multiple plots
library(distributional)
library(ggdist)

# Set ggplot2 theme
black_theme <- theme(
  axis.line = element_line(linewidth = 2, lineend = "round", color = "white"),
  panel.grid = element_blank(),
  panel.background = element_rect(fill = "black", color = NA),
  axis.ticks = element_blank(),
  axis.text = element_text(size = 21, face = "bold", color = "white"),
  axis.title = element_text(size = 21, face = "bold", color = "white"),
  plot.title = element_text(size = 25, face = "bold", color = "white"),
  plot.background = element_rect(fill = "black", color = NA),
  legend.background = element_rect(fill = "black", color = NA),
  legend.text = element_text(size = 21, color = "white"),
  legend.title = element_text(size = 21, face = "bold", color = "white"),
  legend.position = "bottom"
)

theme_set(black_theme)

# Set global default for geom_point shape
update_geom_defaults("point", list(shape = 1, size = 3, alpha = 0.8))

# silence marginaleffects warnings about glmmTMB
options(marginaleffects_safe = FALSE)

set.seed(333)
```

# Case Study: Numerical × Numerical Interactions
BIT ABOUT THE MATH FOR INTERACTIONS

For the simulated examples in this tutorial, we will be measuring the petal length of flowers in a forest. We are interested in the effect of environmental conditions on petal length.

## Simulating the dataset
Let's simulate an initial dataset of 200 petal lengths, temperature and rain measurements.
```{r}
n_flowers <- 200

temperature <- rnorm(n = n_flowers, mean = 10, sd = 15)

rain <- rgamma(n = n_flowers, shape = 2)

avg_petal_length <- 25.6

sd_petal_length <- 4.5

temperature_effect <- 0.4

rain_effect <- 0.9

interaction_effect <- 0.2

tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(
    n = n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered +
      interaction_effect * temp_centered * rain_centered,
    sd = sd_petal_length
  )
) -> flower_measurements_1
```

You may notice in the code above that, in contrast to how we simulated Bighorn Sheep data (LINK to first tutorial), we have centered the values of the predictors. In other words, we have substracted the average temperature or rain value from the observations. This has for effect of shifting the model intercept. Rather than the intercept being the average petal length when rain and temperature = 0, it represents the average petal length for flowers exposed to average temperature and rain conditions. This makes its value (debatably) more pertinent to our research topic.

```{r}
library(GGally)
ggpairs(
  flower_measurements_1,
  diag  = list(continuous = wrap("densityDiag", fill = "white", alpha = 0.6, color = "white")),
  lower = list(continuous = wrap("points", color = "white", alpha = 0.6, size = 2, shape = 1)),
  upper = list(continuous = wrap("cor", color = "white", size = 5))
) +
  theme(
    strip.text = element_text(color = "black", face = "bold", size = 12)
  )
```

Let's fit a linear model to these data.
```{r}
mod_flower_measurements_1 <- glmmTMB(petal_length ~ temp_centered * rain_centered, data = flower_measurements_1)

summary(mod_flower_measurements_1)
```

We can see from the model output that the model estimates our parameters pretty well.

Notice how we have gone straight to a model without having clearly defined our research questions. We now have a model output; every parameter has 3 stars which means we are significantly happy. But why? And, most importantly, what do these values mean?

Since our vague research objective is to quantify the effect of environmental conditions on petal length, let's break down our model and extract as many effects as possible. This will help us understand interactions better and formulate more specific research questions.

Let us start with the Estimates for temp_centered and rain_centered. These are the slopes for temperature and rain when the other variable is set to zero. As we mean-centered these predictors, the estimates are the effect of one variable when the other is at its mean. Let's validate this with the slopes() function.

Temperature effect when rain_centered = 0
```{r}
slopes(mod_flower_measurements_1,
  newdata = datagrid(
    rain_centered = 0
  ),
  variables = "temp_centered"
)
```

Rain effect when temperature_centered = 0
```{r}
slopes(mod_flower_measurements_1,
  newdata = datagrid(
    temp_centered = 0
  ),
  variables = "rain_centered"
)
```

However, these are the effect of one predictor when the other is set to a single value. As we specified an interaction in the model, we may be more interested in the slope of one of the variables across the range of the values of the other. We will add the average back to the centered values to get the raw values of the predictors back.

Effect of rain:
```{r}
slopes_by_temp <- slopes(mod_flower_measurements_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
      max(flower_measurements_1$temp_centered),
      length.out = 30
    )
  ),
  variables = "rain_centered",
  by = "temp_centered"
)

trend_slope_temp <- coef(lm(estimate ~ temp_centered, data = slopes_by_temp))[["temp_centered"]]

slopes_by_temp |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linewidth = 1.5, linetype = 2) +
  geom_ribbon(aes(x = temp_centered + mean(temperature), ymin = conf.low, ymax = conf.high),
    alpha = 0.4,
    fill = "lightgrey"
  ) +
  geom_line(aes(x = temp_centered + mean(temperature), y = estimate),
    lineend = "round",
    linewidth = 2,
    color = "white"
  ) +
  annotate("text",
    x = -Inf, y = Inf,
    hjust = -0.2, vjust = 1.5,
    label = paste0("slope = ", round(trend_slope_temp, 3)),
    color = "white", size = 6, fontface = "bold"
  ) +
  scale_x_continuous(breaks = seq(-25, 50, by = 5)) +
  labs(
    y = "Effect of rain on petal length",
    x = "Temperature values"
  )
```

We can see that rain has a negative effect on petal length when temperatures are below 5.

Effect of temperature:
```{r}
slopes_by_rain <- slopes(mod_flower_measurements_1,
  newdata = datagrid(rain_centered = seq(min(flower_measurements_1$rain_centered),
    max(flower_measurements_1$rain_centered),
    length.out = 30
  )),
  variables = "temp_centered",
  by = "rain_centered"
)

trend_slope <- coef(lm(estimate ~ rain_centered, data = slopes_by_rain))[["rain_centered"]]

slopes_by_rain |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linewidth = 1.5, linetype = 2) +
  geom_ribbon(aes(x = rain_centered + mean(rain), ymin = conf.low, ymax = conf.high),
    alpha = 0.4,
    fill = "lightgrey"
  ) +
  geom_line(aes(x = rain_centered + mean(rain), y = estimate),
    lineend = "round",
    linewidth = 2,
    color = "white"
  ) +
  annotate("text",
    x = -Inf, y = Inf,
    hjust = -0.2, vjust = 1.5,
    label = paste0("slope = ", round(trend_slope, 3)),
    color = "white", size = 6, fontface = "bold"
  ) +
  labs(
    y = "Effect of temperature on petal length",
    x = "Rain values"
  )
```

Temperature has a positive effect for all values of rain but that effect increases as rain increases.

We can obtain similar plots using the plot_slopes() function https://marginaleffects.com/man/r/plot_slopes.html

The slope obtained between the effect of one predictor and the values of the other is the interaction parameter we obtain in the model output. This is the *moderating* effect that the predictors have on one another. As we did not have a very specific research question, we can report both of the plots above as a description of how temperature and rain moderate each other's effect on petal length. In many contexts, however, we have apriori hypotheses about how variables interact within a system. For example, if we study a drought-resistant plant, we may hypothesize that temperature (the *exposure* variable) is the most relevant factor that could affect petal length (the *outcome*) but its effect is moderated by rain (the *moderator*). If that is the case, let's look at some ways to display the model predictions.

We can start by looking at the predicted average petal length as a function of temperature for different values of rain. The choice of moderator values for which we generate predictions can vary depending on the research question. It is common to choose certain quantiles of the moderator. In this example, we will use the five-number summary (https://en.wikipedia.org/wiki/Five-number_summary) for rain values.

```{r}
predictions(mod_flower_measurements_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
      max(flower_measurements_1$temp_centered),
      length.out = 30
    ),
    rain_centered = fivenum
  )
) |>
  as_tibble() |>
  mutate(rain = round(rain_centered + mean(rain), 2)) |>
  ggplot() +
  geom_line(aes(y = estimate,
                x = temp_centered + mean(temperature),
                color = as.factor(rain)),
            lineend = "round",
            linewidth = 1.5) +
  geom_ribbon(aes(ymin = conf.low,
                  ymax = conf.high,
                  x = temp_centered + mean(temperature),
                  fill = as.factor(rain)),
              alpha = 0.2) +
  labs(
    y = "Predicted petal length",
    x = "Temperature",
    color = "Rain",
    fill = "Rain"
  ) +
  coord_cartesian(ylim = c(0, 50))
```

If we ignore the fact that the model predicts negative petal lengths when temperature is low and rain is high (7.79), visualizing the model predictions this way clearly shows how the effect of temperature on petal length varies with/is moderated by rain values.

Another way to visualize these trends is to generate a response surface.

```{r}
predictions(mod_flower_measurements_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
      max(flower_measurements_1$temp_centered),
      length.out = 30
    ),
    rain_centered = seq(min(flower_measurements_1$rain_centered),
      max(flower_measurements_1$rain_centered),
      length.out = 30
    )
  )
) |>
  as_tibble() |>
  mutate(rain = rain_centered + mean(rain),
         temp = temp_centered + mean(temperature)) |>
  ggplot(aes(x = rain, y = temp, z = estimate)) +
  stat_contour_filled() +
  labs(
    y = "Temperature",
    x = "Rain",
    fill = "Predicted petal length"
  )
```

The advantage of these plots is that we can see how the predictions change for all combinations of the exposure (temperature) and moderator (rain). For example, if we focus on rain = 6 and look up from the bottom of the y-axis to the top (as temperature increases), we can see the predicted petal length increase substantially (again, ignoring the fact that negative values are predicted). I personally prefer plots such as the one with predictions for the five-number summary as they are less of a mind-bender and it is easier to see the uncertainty related to the predicted mean. I've also noticed that response surfaces displayed as such lead people to over-interpretation of the results. As we can arbitrarily change the values in each color bin (yellow could be for predicted petal lengths of 60-90 instead of 80-90), and the 95% confidence intervals are missing, we can overemphasize effects that, in reality, are poorly supported by the model.

For fun, let's visualize this response surface in 3D with the 95% confidence intervals. I would not necessarily recommend these plots for a publication but they can be a useful way to better understand the model predictions and their uncertainty. These plots will also be informative for future topics in this tutorial series.

```{r}
#| warning: false
library(ggcube)

predictions(mod_flower_measurements_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
      max(flower_measurements_1$temp_centered),
      length.out = 30
    ),
    rain_centered = seq(min(flower_measurements_1$rain_centered),
      max(flower_measurements_1$rain_centered),
      length.out = 30
    )
  )
) |>
  as_tibble() |>
  mutate(
    rain = rain_centered + mean(rain),
    temp = temp_centered + mean(temperature)
  ) |>
  ggplot() +
  stat_surface_3d(aes(x = rain, y = temp, z = conf.high), fill = "red", alpha = 0.4) +
  stat_surface_3d(aes(x = rain, y = temp, z = estimate), fill = "blue") +
  stat_surface_3d(aes(x = rain, y = temp, z = conf.low), fill = "red", alpha = 0.4) +
  coord_3d(light = "none") +
  labs(
    z = "Petal length",
    x = "Rain",
    y = "Temp"
  ) -> resp_surface_3d

#animate_3d(resp_surface_3d, yaw = c(0, 360))
```

The animation does not render the way I would like it to but I hope the reader can still appreciate the shape of the response surface in all its dimensions. We will explore response surfaces of different shapes in a future tutorial on interactions with Generalized Additive Models (GAMs).

# Bonus Case Study: Variance-moderating effects
Let's push our petal length example a bit further and define our research question more clearly. Previously, we had been interested in how rain moderates the effect of temperature on the *average* petal length. Suppose now that we have individuals in our population that are cold-resistant whereas other individuals are less resistant to the point that when temperatures are below zero, these non-cold-resistant individuals die before we get to measure them. Under this scenario, temperature resistance genes are linked to petal growth such that a reduction in genetic diversity at low temperatures leads to a reduction in the amplitude of the response of petal length to low temperatures at the population level. In other words, a reduction in genetic diversity reduces the *variance* of observed population petal length at low temperatures. As temperature increases, the average petal length responds in the same way as in the previous examples but the variance of these observations increases. For now, we are leaving rain aside to concentrate on a simpler example of how we can analyse data that are generated from this scenario.

It turns out that the described scenario leads to heteroscedasticity around the average petal length as a function of temperature; the variance is not constant around the average through the trend. It turns out that we can use glmmTMB to take into account this heteroscedasticity by modeling the standard deviation of petal length. These types of regressions are called *distributional regression* because we model more than one parameter of the response distribution (in this case, we want to model the mean and standard deviation of the Normal distribution). glmmTMB estimates the standard deviation on the log scale which assumes an exponential increase in the observed standard deviation (or variance).

Let's simulate data.

```{r}
# Dispersion model parameters (on log scale to match glmmTMB)
log_sd_intercept <- log(sd_petal_length)  # log(sd) when temp_centered = 0
log_sd_temp_slope <- 0.03  # Effect of temp on log(sd)

tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(
    n = n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered,
    sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered)
  )
) -> flower_measurements_2


ggplot(flower_measurements_2, aes(x = temp_centered + mean(temperature), y = petal_length)) +
  geom_smooth(method = 'lm',     
              lineend = "round",
              linewidth = 2,
              color = "white",
              se = F) + 
  geom_point(color = "white") +
  labs(x = "Temperature",
       y = "Observed petal length")
```

We can see here that the chosen parameter values reflect exactly our scenario - the observed petal length variance (standard deviation squared) increases starting around temperatures of zero.

We can model the standard deviation using the dispformula argument in glmmTMB

```{r}
mod_flower_measurements_2 <- glmmTMB(petal_length ~ temp_centered,
                                     dispformula = ~ temp_centered,
                                     data = flower_measurements_2)

summary(mod_flower_measurements_2)

# Check simulated sd parameters
cat("\nExpected dispersion intercept (log scale):", log_sd_intercept, "\n")
cat("Expected dispersion slope:", log_sd_temp_slope, "\n")
```

The model recovers our simulated parameters pretty well

Let's visualize how well the model recovers the variance structure:
We do this by predicting the standard deviation values for all observed values of temperature via the type = "disp" argument in predictions()
```{r}
# estimate sd
predictions(mod_flower_measurements_2, newdata = flower_measurements_2, type = "disp") |> as_tibble() -> pred_sd_temp

# simulated sd
flower_measurements_2 |>
  mutate(
    true_sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered)
  ) -> true_sd_temp

# Plot comparison
ggplot() +
  # True simulated variance
  geom_line(data = true_sd_temp, 
            aes(x = temp_centered + mean(temperature), 
                y = true_sd, color = "Simulated"),
            linewidth = 2, 
            linetype = "dashed"
  ) +

  geom_ribbon(data = pred_sd_temp, 
              aes(ymin = conf.low, ymax = conf.high, 
                  x = temp_centered + mean(temperature)), 
              fill = "red", 
              alpha = 0.6) +

  geom_line(data = pred_sd_temp, 
            aes(x = temp_centered + mean(temperature), 
                y = estimate, 
                color = "Estimated"),
            linewidth = 2,
            lineend = "round"
  ) +

  scale_color_manual(
    values = c("Simulated" = "green", "Estimated" = "red"),
    name = "Standard deviation"
  ) +
  labs(
    x = "Temperature",
    y = "Standard deviation"
  )
```

The model recovers the simulated relationship between petal standard deviation and temperature pretty well.

Now that we understand distributional regression a bit better, let's look at moderating effects for the standard deviation (or variance) and how we can communicate the results in the most interpretable ways. We will get funky here and simulate only simple effect of temperature and rain on the mean petal length (without interaction) and add an interaction between the two variables on the standard deviation of petal length. Distributional regression is essentially a combination of (linear) models for each of the modeled distribution parameters so it may actually be more useful in certain cases to have different model formulas (and/or predictors) for the different distribution parameters, depending on the research question.

```{r}
log_sd_rain_slope <- 0.02         # effect of rain on log(sd)
log_sd_interaction_effect <- 0.005 # interacting effect of temp and rain on log(sd)

tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(
    n = n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered,
    sd = exp(log_sd_intercept + 
      log_sd_temp_slope * temp_centered +
      log_sd_rain_slope * rain_centered +
      log_sd_interaction_effect* temp_centered * rain_centered)
  )
) -> flower_measurements_3


ggpairs(
  flower_measurements_3,
  diag  = list(continuous = wrap("densityDiag", fill = "white", alpha = 0.6, color = "white")),
  lower = list(continuous = wrap("points", color = "white", alpha = 0.6, size = 2, shape = 1)),
  upper = list(continuous = wrap("cor", color = "white", size = 5))
) +
  theme(
    strip.text = element_text(color = "black", face = "bold", size = 12)
  )
```

The model

```{r}
mod_flower_measurements_3 <- glmmTMB(petal_length ~ temp_centered + rain_centered,
                                     dispformula = ~ temp_centered+ rain_centered + temp_centered*rain_centered,
                                     data = flower_measurements_3)

summary(mod_flower_measurements_3)

# Check simulated sd parameters
cat("\nExpected dispersion intercept (log scale):", log_sd_intercept, "\n")
cat("Expected dispersion temperature slope:", log_sd_temp_slope, "\n")
cat("Expected dispersion rain slope:", log_sd_rain_slope, "\n")
cat("Expected dispersion interaction slope:", log_sd_interaction_effect, "\n")
```

Let's go through an entire workflow for reporting all of the results we generated with this model.

First, let's start with the predictors' effects on the average petal length. We were not interested in an interaction between these two variables for the average so we will plot the predicted average of petal as a function of the two predictors side-by-side.

```{r}
#Temperature effect when rain_centered = 0
predictions(mod_flower_measurements_3,
  newdata = datagrid(
    rain_centered = 0
  ),
  variables = "temp_centered"
) |> as_tibble() -> pred_temp

ggplot(pred_temp) +
  geom_ribbon(
              aes(ymin = conf.low, ymax = conf.high, 
                  x = temp_centered + mean(temperature)), 
              fill = "white", 
              alpha = 0.2) +
  geom_line( 
            aes(x = temp_centered + mean(temperature), 
                y = estimate),
            linewidth = 2,
            lineend = "round",
            color = "white") +
  labs(y = "Predicted average petal length",
       x = "Temperature") -> temp_avg_plot

# Rain effect when temp_centered = 0
predictions(mod_flower_measurements_3,
  newdata = datagrid(
    temp_centered = 0
  ),
  variables = "rain_centered"
) |> as_tibble() -> pred_rain

ggplot(pred_rain) +
  geom_ribbon(
              aes(ymin = conf.low, ymax = conf.high, 
                  x = rain_centered + mean(rain)), 
              fill = "white", 
              alpha = 0.2) +
  geom_line( 
            aes(x = rain_centered + mean(rain), 
                y = estimate),
            linewidth = 2,
            lineend = "round",
            color = "white") +
  labs(y = "Predicted average petal length",
       x = "Rain") -> rain_avg_plot

temp_avg_plot | rain_avg_plot
```

These are the *conditional* effects of each of the two predictors on the average petal length.

Now let's look at the interacting effect of temperature and rain on the standard deviation in the ways we did as for the average petal length in the first example.

```{r}
predictions(mod_flower_measurements_3,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_3$temp_centered),
      max(flower_measurements_3$temp_centered),
      length.out = 30
    ),
    rain_centered = fivenum
  ),
  type = "disp"
) |>
  as_tibble() |>
  mutate(rain = round(rain_centered + mean(rain), 2)) |>
  ggplot() +
  geom_line(aes(y = estimate,
                x = temp_centered + mean(temperature),
                color = as.factor(rain)),
            lineend = "round",
            linewidth = 1.5) +
  geom_ribbon(aes(ymin = conf.low,
                  ymax = conf.high,
                  x = temp_centered + mean(temperature),
                  fill = as.factor(rain)),
              alpha = 0.2) +
  labs(
    y = "Predicted standard deviation of petal length",
    x = "Temperature",
    color = "Rain",
    fill = "Rain"
  ) +
  coord_cartesian(ylim = c(0, 50))
```

Now, with a response surface

```{r}
predictions(mod_flower_measurements_3,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_3$temp_centered),
      max(flower_measurements_3$temp_centered),
      length.out = 30
    ),
    rain_centered = seq(min(flower_measurements_3$rain_centered),
      max(flower_measurements_3$rain_centered),
      length.out = 30
    )
  ),
  type = "disp"
) |>
  as_tibble() |>
  mutate(rain = rain_centered + mean(rain),
         temp = temp_centered + mean(temperature)) |>
  ggplot(aes(x = rain, y = temp, z = estimate)) +
  stat_contour_filled() +
  labs(
    y = "Temperature",
    x = "Rain",
    fill = "Predicted sd of petal length"
  )
```

The response surface is not great in this specific example because extremely high standard deviations are predicted when temperature and rain are high.

We can also look at the mediating effect of temperature on the effect of rain on petal standard deviation

```{r}
slopes_by_temp_sd <- slopes(mod_flower_measurements_3,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_3$temp_centered),
      max(flower_measurements_3$temp_centered),
      length.out = 30
    )
  ),
  variables = "rain_centered",
  by = "temp_centered",
  type = "disp"
)

slopes_by_temp_sd |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linewidth = 1.5, linetype = 2) +
  geom_ribbon(aes(x = temp_centered + mean(temperature), ymin = conf.low, ymax = conf.high),
    alpha = 0.4,
    fill = "lightgrey"
  ) +
  geom_line(aes(x = temp_centered + mean(temperature), y = estimate),
    lineend = "round",
    linewidth = 2,
    color = "white"
  ) +
  scale_x_continuous(breaks = seq(-25, 50, by = 5)) +
  labs(
    y = "Effect of rain on sd petal length",
    x = "Temperature values"
  )
```

And the inverse, the mediating effect of rain on the effect of temperature on petal standard deviation

```{r}
slopes_by_rain_sd <- slopes(mod_flower_measurements_3,
  newdata = datagrid(rain_centered = seq(min(flower_measurements_3$rain_centered),
    max(flower_measurements_3$rain_centered),
    length.out = 30
  )),
  variables = "temp_centered",
  by = "rain_centered",
  type = "disp"
)

slopes_by_rain_sd |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linewidth = 1.5, linetype = 2) +
  geom_ribbon(aes(x = rain_centered + mean(rain), ymin = conf.low, ymax = conf.high),
    alpha = 0.4,
    fill = "lightgrey"
  ) +
  geom_line(aes(x = rain_centered + mean(rain), y = estimate),
    lineend = "round",
    linewidth = 2,
    color = "white"
  ) +
  labs(
    y = "Effect of temperature on sd petal length",
    x = "Rain values"
  )
```

While reporting the predictors' effects on the average and standard deviation of petal length separately is informative, I personally find it difficult to interpret knowing that the two parameters of the Normal distribution act together to generate the observations. I suggest in this case to also create figures that communicate the predicted observations generated from the estimate values of the average and standard deviation. We can do so by predicting the average and standard deviation based on different values of temperature and rain. In the following code, we predict the average petal length using the type = "conditional" argument in predictions() and join these values to the predicted standard deviation using the type = "disp" argument. 
We then use the distributional package which I absolutely love to generate distributions. In this specific case, we use the dist_normal() function to which we give the predicted values of the average and standard deviation of petal length for every combination of predictor values. The result gives an entire distribution contained within a single cell in the dataset. In other words, every row of the pred_obs column is a distribution of predicted observations of petal length.

```{r}
predictions(mod_flower_measurements_3,
            newdata = datagrid(
              temp_centered = fivenum,
              rain_centered = fivenum
            ),
            type = "conditional") |>
  as_tibble() |>
  select(temp_centered, rain_centered, estimate) |>
  rename(mu = estimate) |>
  left_join(predictions(mod_flower_measurements_3,
            newdata = datagrid(
              temp_centered = fivenum,
              rain_centered = fivenum
            ),
            type = "disp") |>
  as_tibble() |>
  select(temp_centered, rain_centered, estimate) |>
  rename(sd = estimate)) |>
  rowwise() |>
  mutate(temp_centered = round(temp_centered),
         rain_centered = round(rain_centered),
         pred_obs = dist_normal(mu = mu, sd = sd)) -> predicted_observations

temp_labels <- c(
  `-38` = "temp = -38",
  `-9` = "temp = -9",
  `1` = "temp = 1",
  `9` = "temp = 9",
  `36` = "temp = 36")

ggplot(predicted_observations, aes(xdist = pred_obs, color = as.factor(rain_centered))) +
  stat_slab(fill = NA, normalize = "xy") +
  scale_color_brewer(palette = "Accent") +
  facet_wrap(~ as.factor(temp_centered), scales = "free", labeller = as_labeller(temp_labels)) +
  labs(x = "Predicted petal length", color = "Rain (centered)") +
  theme(axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(color = "black", face = "bold", size = 12))
```

This type of figure provides a richer understanding of the model, in my opinion. We can see that petal length is generally similar across values of temperature and rain except when rain is in extremely high values.