---
title: "Understanding Interactions in Regression Models"
subtitle: "Categorical × numerical and numerical × numerical interactions, marginal vs. conditional effects, and common pitfalls in ecological data"
author: "Allen Bush-Beaupré"
date: 10-05-2026
categories: [tutorial, interactions, marginaleffects, glmm, statistics, R, workshop]
format:
  revealjs:
    theme: black
    slide-number: true
    chalkboard: true
    code-line-numbers: true
    fig-width: 9
    fig-height: 6
    out-width: "90%"
    footer: "Interactions in Regression Models"
    css: interactions_regression.css
execute:
  echo: true
  warning: false
---

## Welcome

This workshop is adapted from a two-part tutorial series on interactions in regression models:

- Part 1: [Categorical × numerical interactions](../../posts/interactions_categorical_numerical/index.qmd)
- Part 2: [Numerical × numerical interactions](../../posts/interactions_numerical_numerical/index.qmd)

**Audience**: graduate students and researchers in ecology who fit GL(M)Ms and want to interpret interaction terms correctly.

**Format**: short lecture blocks, live-coded examples, and practice exercises (look for the 🧪 slides).

## Learning objectives

By the end of this workshop you should be able to:

1.  Explain what an interaction term means for a categorical × numerical model and for a numerical × numerical model
2.  Distinguish **conditional** effects from **marginal** effects, and choose the right one for a research question
3.  Recognize Simpson's Paradox and sampling bias as threats to naive interpretation
4.  Use `marginaleffects` (`predictions()`, `slopes()`, `avg_predictions()`, `avg_slopes()`, `hypotheses()`) to extract and test the quantities you actually care about
5.  Visualize interaction effects (slope-by-moderator plots, five-number-summary predictions, response surfaces) and know their trade-offs

## Setup

```{r}
#| output: false
library(tidyverse)       # Data manipulation and plotting
library(glmmTMB)         # Generalized linear (mixed) models
library(marginaleffects) # Compute marginal/conditional effects
library(patchwork)       # Combine multiple plots
library(GGally)          # Pairwise plots

# Set ggplot2 theme
black_theme <- theme(
  axis.line = element_line(linewidth = 2, lineend = "round", color = "white"),
  panel.grid = element_blank(),
  panel.background = element_rect(fill = "black", color = NA),
  axis.ticks = element_blank(),
  axis.text = element_text(size = 21, face = "bold", color = "white"),
  axis.title = element_text(size = 21, face = "bold", color = "white"),
  plot.title = element_text(size = 25, face = "bold", color = "white"),
  plot.background = element_rect(fill = "black", color = NA),
  legend.background = element_rect(fill = "black", color = NA),
  legend.text = element_text(size = 21, color = "white"),
  legend.title = element_text(size = 21, face = "bold", color = "white"),
  legend.position = "bottom"
)

theme_set(black_theme)

# Set global default for geom_point shape
update_geom_defaults("point", list(shape = 1, size = 3, alpha = 0.8))

# silence marginaleffects warnings about glmmTMB
options(marginaleffects_safe = FALSE)

set.seed(4127)
```

# Part 1 — Categorical × Numerical Interactions

## What is an interaction?

For predictors $X_1$ and $X_2$:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 (X_1 \times X_2) + \epsilon$$

$\beta_3$ is the interaction coefficient. When $\beta_3 \neq 0$, the effect of $X_1$ on $Y$ **depends on the value of** $X_2$.

For a categorical $X_2$ (e.g., sex) and continuous $X_1$ (e.g., snow depth): the interaction means the **slope** of the continuous predictor differs between categories.

## Conditional vs. marginal effects

- **Conditional effects**: the effect of a predictor *at specific values* of other predictors (e.g., the slope of snow depth *for males*)
- **Marginal effects**: the effect *averaged across* the distribution of other predictors (e.g., the *average* slope of snow depth across the whole population)

Both are legitimate; they answer different questions. Interactions are exactly where the distinction starts to matter.

## Case study: Bighorn sheep

Ram Mountain, Alberta. Question: how does winter snow depth relate to spring mass in Bighorn sheep, and does it differ by sex? Males and females segregate in winter — females use lower, less snowy habitat.

```{r}
n_individuals_population <- 1000
sexes <- c("male", "female")
male_intercept <- 220; female_intercept <- 180
sd_weight <- 15
male_loss <- -0.8; female_loss <- -2.5

bighorn_population <- tibble(
  sex = rep(sexes, n_individuals_population / length(sexes)),
  snow_depth = runif(n_individuals_population,
    min = ifelse(sex == "male", 35, 0),
    max = ifelse(sex == "male", 70, 35)
  ),
  weight = rnorm(n_individuals_population,
    mean = ifelse(sex == "male",
      male_intercept + male_loss * snow_depth,
      female_intercept + female_loss * snow_depth),
    sd = sd_weight
  )
)
```

Built-in features: different intercepts, different slopes, **and** different snow-depth distributions by sex.

## Simpson's Paradox {.smaller .fig-slide}

```{r}
#| code-fold: true
#| fig-cap: "Ignoring sex hides the true within-sex trends."
ggplot(bighorn_population, aes(x = snow_depth, y = weight)) +
  geom_point(aes(color = sex)) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 2)
```

Pooled: snow depth looks **positively** associated with weight — even though both sexes lose weight with more snow.

## Why? Sex is a confounder

- Within each sex: more snow → lower weight (true negative relationship)
- Between sexes: males see more snow **and** are heavier (confounded)
- Pooled data compares heavy, high-snow males to light, low-snow females → spurious positive trend

## Within-sex trends {.smaller .fig-slide}

```{r}
#| code-fold: true
#| fig-cap: "Within-sex trends are negative, as expected."
ggplot(bighorn_population, aes(x = snow_depth, y = weight, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## 🧪 Exercise 1

Using `bighorn_population`:

1.  Fit `weight ~ sex * snow_depth` with `glmmTMB()`. Call it `mod_mass_true`.
2.  From the summary, identify which coefficient represents the **male** slope of snow depth. (Hint: it's not printed directly — you have to add two rows together.)
3.  In one sentence, explain why a naive `weight ~ snow_depth` model (ignoring sex) would mislead a manager trying to understand the effect of a harsh winter.

*(\~5 minutes — discuss with a neighbor)*

## Sampling from the population

In practice we can't measure everyone. Two sampling scenarios:

- **Scenario 1**: balanced trap, 50 males / 50 females
- **Scenario 2**: biased trap, 80 males / 20 females

```{r}
n_individuals <- 100
bighorn_mass_1 <- bighorn_population |>
  group_by(sex) |>
  slice_sample(n = n_individuals / 2, replace = FALSE) |>
  ungroup()

mod_mass_1 <- glmmTMB(weight ~ sex * snow_depth,
                       data = bighorn_mass_1, family = gaussian())
```

## Reading the coefficients

- **Intercept**: female weight at snow_depth = 0
- **sexmale**: male − female intercept difference
- **snow_depth**: female slope
- **sexmale:snow_depth**: male − female slope difference

Useful for hypothesis tests, not directly interpretable for "what's the effect for males?" — that's what `marginaleffects` is for.

## Model coefficients {.smaller}

```{r}
summary(mod_mass_1)$coefficients$cond
```

## Q1: Do sexes differ in spring mass?

```{r}
predictions(mod_mass_1, by = "sex")
```

This is a **marginal prediction**: the model-implied mean for each sex, averaged over the observed snow-depth distribution in the sample.

```{r}
predictions(mod_mass_1, by = "sex") |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

## Q2: Does snow depth affect mass equally by sex? {.smaller .fig-slide}

```{r}
#| code-fold: true
#| fig-cap: "Slopes of snow depth by sex, with true simulated values as vertical lines."
slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  ggplot(aes(x = estimate, xmax = conf.high, xmin = conf.low, y = sex, color = sex)) +
  geom_pointrange(size = 1.5, lineend = "round", linewidth = 2, pch = 21, stroke = 4) +
  geom_vline(xintercept = female_loss, color = "red") +
  geom_vline(xintercept = male_loss, color = "blue") +
  theme(legend.position = "none")
```

## Testing the slope difference

```{r}
slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

## Q3: What's the population-average slope?

`avg_slopes()` computes the slope for every observation, then averages:

```{r}
avg_slopes(mod_mass_1, variables = "snow_depth", vcov = TRUE)
```

```{r}
mean(c(male_loss, female_loss))  # true simulated average
```

## Marginal vs conditional, visualized {.smaller .fig-slide}

```{r}
#| code-fold: true
avg_pred_mass_1 <- avg_predictions(mod_mass_1, variables = "snow_depth")
pred_sex_mass_1 <- predictions(mod_mass_1, newdata = bighorn_mass_1)

ggplot() +
  geom_ribbon(data = avg_pred_mass_1,
              aes(x = snow_depth, ymin = conf.low, ymax = conf.high),
              color = "lightgreen", fill = "lightgreen", alpha = 0.4) +
  geom_line(data = avg_pred_mass_1, aes(x = snow_depth, y = estimate),
            color = "lightgreen", lineend = "round", linewidth = 2) +
  geom_ribbon(data = pred_sex_mass_1,
              aes(x = snow_depth, ymin = conf.low, ymax = conf.high, color = sex, fill = sex),
              alpha = 0.8) +
  geom_line(data = pred_sex_mass_1, aes(x = snow_depth, y = estimate, color = sex),
            lineend = "round", linewidth = 2) +
  geom_point(data = bighorn_mass_1, aes(y = weight, x = snow_depth, color = sex)) +
  guides(alpha = "none", fill = "none") +
  labs(title = "Green = marginal; red/blue = conditional (by sex)") +
  coord_cartesian(ylim = c(50, 220))
```

## Scenario 2: biased sampling (80% male)

```{r}
bighorn_mass_2 <- bind_rows(
  bighorn_population |> filter(sex == "male") |> slice_sample(n = 80),
  bighorn_population |> filter(sex == "female") |> slice_sample(n = 20)
)
mod_mass_2 <- glmmTMB(weight ~ sex * snow_depth, data = bighorn_mass_2, family = gaussian())
avg_slopes(mod_mass_2, variables = "snow_depth", vcov = TRUE)
```

Conditional (by-sex) slopes stay close to truth. But the **marginal** slope is biased toward the overrepresented, weaker-effect sex (males).

## Correcting for known sampling bias

If we *know* the true population sex ratio (50:50), we can average over a **balanced** `newdata` grid instead of our biased sample.

**Key principle**: model coefficients capture conditional relationships from your sample; marginal effects can target *any* population composition via `newdata`.

## Balanced-`newdata` correction {.smaller}

```{r}
newdata_balanced <- bighorn_mass_2 |>
  group_by(sex) |>
  reframe(snow_depth = seq(min(snow_depth), max(snow_depth), length.out = 50)) |>
  ungroup()

avg_predictions(mod_mass_2, variables = "snow_depth", newdata = newdata_balanced) |>
  as_tibble() |> select(estimate, conf.low, conf.high) |> head(3)
```

## 🧪 Exercise 2

Using `mod_mass_2` (the biased sample model):

1.  Compute `avg_slopes()` **without** correcting for sampling bias, and again **with** the balanced `newdata_balanced` grid.
2.  Which one is closer to the true average slope, `mean(c(male_loss, female_loss))`?
3.  Suppose instead you *didn't* know the true sex ratio in the population — what additional data would you need to collect to justify a correction like this?

*(\~7 minutes — try it in your own script, then we'll compare answers)*

## Part 1 takeaways

1.  Interactions → conditional effects that depend on other variables
2.  Simpson's Paradox is a real risk when pooling across confounded groups
3.  Marginal ≠ conditional; both are valid but answer different questions
4.  Raw model coefficients rarely equal your research quantity of interest — use `marginaleffects`
5.  Marginal effects inherit your **sample's** covariate distribution unless you override it with `newdata`

# Part 2 — Numerical × Numerical Interactions

## The math, revisited

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 (X_1 \times X_2) + \epsilon$$

Regrouping terms in $X_1$:

$$Y = \beta_0 + (\beta_1 + \beta_3 X_2)\, X_1 + \beta_2 X_2 + \epsilon$$

The slope of $X_1$ is now **itself a linear function of** $X_2$: it equals $\beta_1 + \beta_3 X_2$. Symmetrically, the slope of $X_2$ is $\beta_2 + \beta_3 X_1$.

$\beta_1$ alone is only the slope of $X_1$ when $X_2 = 0$ — meaningful only if zero is a meaningful value of $X_2$. This motivates **centering**.

## Case study: petal length

Simulated flowers; petal length as a function of temperature and rain, with an interaction.

Predictors are **mean-centered**: the intercept becomes the predicted response at *average* conditions rather than at (biologically meaningless) zero temperature/rain.

## Simulating the data {.smaller}

```{r}
n_flowers <- 200
temperature <- rnorm(n_flowers, mean = 10, sd = 15)
rain <- rgamma(n_flowers, shape = 2)
avg_petal_length <- 25.6; sd_petal_length <- 4.5
temperature_effect <- 0.4; rain_effect <- 0.9; interaction_effect <- 0.2

flower_measurements_1 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered +
      interaction_effect * temp_centered * rain_centered,
    sd = sd_petal_length)
)
```

## Fit the model

```{r}
mod_flower_1 <- glmmTMB(petal_length ~ temp_centered * rain_centered,
                         data = flower_measurements_1)
summary(mod_flower_1)$coefficients$cond
```

Every term has a small p-value — but what do these numbers *mean* for our research question? Time to unpack them.

## Slope of one predictor at a single value of the other

```{r}
# Effect of temperature when rain_centered = 0 (i.e., average rain)
slopes(mod_flower_1, newdata = datagrid(rain_centered = 0),
       variables = "temp_centered")
```

This matches the `temp_centered` coefficient directly — because rain is centered, "at rain = 0" means "at average rain."

## Slope across the *range* of the moderator {.smaller .fig-slide}

```{r}
#| code-fold: true
#| fig-cap: "Effect of rain on petal length, across observed temperatures."
slopes_by_temp <- slopes(mod_flower_1,
  newdata = datagrid(temp_centered = seq(min(flower_measurements_1$temp_centered),
                                          max(flower_measurements_1$temp_centered), length.out = 30)),
  variables = "rain_centered", by = "temp_centered")

trend_slope_temp <- coef(lm(estimate ~ temp_centered, data = slopes_by_temp))[["temp_centered"]]

slopes_by_temp |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linewidth = 1.5, linetype = 2) +
  geom_ribbon(aes(x = temp_centered + mean(temperature), ymin = conf.low, ymax = conf.high),
    alpha = 0.4, fill = "lightgrey") +
  geom_line(aes(x = temp_centered + mean(temperature), y = estimate),
    lineend = "round", linewidth = 2, color = "white") +
  annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
    label = paste0("slope = ", round(trend_slope_temp, 3)),
    color = "white", size = 6, fontface = "bold") +
  scale_x_continuous(breaks = seq(-25, 50, by = 5)) +
  labs(y = "Effect of rain on petal length", x = "Temperature values")
```

The interaction coefficient is exactly the slope of this line — how much the effect of rain *changes* per unit of temperature.

## Visualizing predictions at chosen moderator values {.smaller .fig-slide}

Common choice: the [five-number summary](https://en.wikipedia.org/wiki/Five-number_summary) of the moderator.

```{r}
#| code-fold: true
predictions(mod_flower_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
                         max(flower_measurements_1$temp_centered), length.out = 30),
    rain_centered = fivenum
  )) |>
  as_tibble() |>
  mutate(rain = round(rain_centered + mean(rain), 2)) |>
  ggplot() +
  geom_line(aes(y = estimate, x = temp_centered + mean(temperature), color = as.factor(rain)),
            lineend = "round", linewidth = 1.5) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high,
                  x = temp_centered + mean(temperature), fill = as.factor(rain)), alpha = 0.2) +
  labs(y = "Predicted petal length", x = "Temperature", color = "Rain", fill = "Rain") +
  coord_cartesian(ylim = c(0, 50))
```

## Response surfaces — use with caution {.smaller .fig-slide}

```{r}
#| code-fold: true
predictions(mod_flower_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
                         max(flower_measurements_1$temp_centered), length.out = 30),
    rain_centered = seq(min(flower_measurements_1$rain_centered),
                         max(flower_measurements_1$rain_centered), length.out = 30)
  )) |>
  as_tibble() |>
  mutate(rain = rain_centered + mean(rain), temp = temp_centered + mean(temperature)) |>
  ggplot(aes(x = rain, y = temp, z = estimate)) +
  stat_contour_filled() +
  labs(x = "Rain", y = "Temperature", fill = "Predicted petal length")
```

Contour bins are arbitrary and hide uncertainty — good for exploration, risky for publication-quality inference. Prefer line-and-ribbon plots when you can.

## 🧪 Exercise 3

Using `mod_flower_1`:

1.  Compute the slope of **temperature** across the observed range of **rain** (mirror the rain-across-temperature plot above).
2.  At what value of rain does the temperature effect become non-significant (CI crosses zero), if at all?
3.  Rewrite the research question this model could answer as a single sentence, identifying which variable is the *exposure* and which is the *moderator*.

*(\~7 minutes)*

## Interactions beyond the mean: distributional regression

Interactions don't have to act only on the average response. `glmmTMB`'s `dispformula` lets predictors affect the **standard deviation** too — useful when variance itself is ecologically meaningful (e.g., reduced trait variance under stress).

## Simulating heteroscedastic data {.smaller}

```{r}
log_sd_intercept <- log(sd_petal_length)
log_sd_temp_slope <- 0.03

flower_measurements_2 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length + temperature_effect * temp_centered,
    sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))
)

mod_flower_2 <- glmmTMB(petal_length ~ temp_centered,
                         dispformula = ~ temp_centered,
                         data = flower_measurements_2)
```

## Checking the dispersion model {.smaller .fig-slide}

```{r}
#| code-fold: true
#| fig-cap: "Recovered vs. simulated standard deviation of petal length."
predictions(mod_flower_2, newdata = flower_measurements_2, type = "disp") |>
  as_tibble() -> pred_sd_temp

true_sd_temp <- flower_measurements_2 |>
  mutate(true_sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))

ggplot() +
  geom_line(data = true_sd_temp,
            aes(x = temp_centered + mean(temperature), y = true_sd, color = "Simulated"),
            linewidth = 2, linetype = "dashed") +
  geom_ribbon(data = pred_sd_temp,
              aes(ymin = conf.low, ymax = conf.high, x = temp_centered + mean(temperature)),
              fill = "red", alpha = 0.6) +
  geom_line(data = pred_sd_temp,
            aes(x = temp_centered + mean(temperature), y = estimate, color = "Estimated"),
            linewidth = 2, lineend = "round") +
  scale_color_manual(values = c("Simulated" = "green", "Estimated" = "red"), name = "Standard deviation") +
  labs(x = "Temperature", y = "Standard deviation")
```

`type = "disp"` retrieves predictions/slopes for the dispersion sub-model, exactly as `type = "conditional"` does for the mean.

## 🧪 Exercise 4 (challenge)

Extend `mod_flower_2` by adding `rain_centered` and a `temp_centered * rain_centered` interaction to **both** the conditional formula and the `dispformula`.

1.  Fit the model on simulated data of your own design (or reuse the logic from `flower_measurements_2`, adding a rain effect and interaction term on the log-sd scale).
2.  Produce one plot showing how the effect of rain on the **standard deviation** of petal length changes across temperature.
3.  In plain language, describe a real ecological scenario (not necessarily plants) where an interaction on variance, rather than the mean, would be the interesting finding.

*(\~10 minutes — this is open-ended; there's no single correct answer)*

## Part 2 takeaways

1.  With two numerical predictors, the interaction coefficient is the *rate of change of one slope per unit of the other* — not a separate slope for each group
2.  Centering predictors makes main-effect coefficients interpretable as "effect at average conditions"
3.  Plot slopes across the moderator's range, not just the coefficient table
4.  Five-number-summary prediction plots are usually more honest than response surfaces (they retain uncertainty)
5.  Interactions can act on variance, not just the mean, via distributional regression (`dispformula`)

## Wrap-up and further reading

- Full write-ups with additional detail and 3D visualizations: [Part 1](../../posts/interactions_categorical_numerical/index.qmd), [Part 2](../../posts/interactions_numerical_numerical/index.qmd)
- `marginaleffects` documentation: <https://marginaleffects.com/>
- Spake et al. (2023), on interactions in GLMs: <https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12939>
- A planned Part 3 will extend this material to non-Gaussian GLMs (non-linear link functions) and Part 4 to GAMs

## Exercise solutions (instructor notes)

::: {.callout-note collapse="true"}
## Exercise 1

- `mod_mass_true <- glmmTMB(weight ~ sex * snow_depth, data = bighorn_population, family = gaussian())`
- Male slope = `snow_depth` coefficient + `sexmale:snow_depth` coefficient
- A naive pooled model would show a *positive* (or attenuated) relationship due to Simpson's Paradox: sex confounds both snow exposure and baseline weight, so ignoring it would lead a manager to underestimate (or misjudge the direction of) the harm of a harsh winter.
:::

::: {.callout-note collapse="true"}
## Exercise 2

- Uncorrected `avg_slopes()` on the biased sample is pulled toward the male slope (weaker effect) because males are overrepresented
- Corrected version (using `newdata_balanced`) should land closer to `mean(c(male_loss, female_loss))`
- Without a known true ratio, you would need an independent, unbiased estimate of population sex ratio (e.g., aerial survey, camera trap monitoring not subject to the same capture bias) to justify reweighting
:::

::: {.callout-note collapse="true"}
## Exercise 3

- Mirror the code from the "slope across the range of the moderator" slide, swapping the roles of `temp_centered` and `rain_centered`
- Answer depends on the random seed / simulated draw — have students report their own CI crossing point
- Example framing: "Does the effect of temperature (exposure) on petal length depend on rainfall (moderator)?"
:::