---
title: "Interactions in Regression Models — Student Handout"
subtitle: "Follow-along script and exercises"
author: "Your name here"
date: 10-05-2026
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: false
---

## Before you start

Run the chunk below **once** to install everything you need. It's written to avoid the most common `glmmTMB` installation failures:

- **Binary vs. source mismatch**: `glmmTMB` links against `TMB` and `Matrix`, which must be built with compatible ABI versions. Installing all three together from the same repository snapshot (rather than upgrading one in isolation) avoids "package was installed with a different version" load-time errors.
- **Stale compiled objects after an R or `Matrix` upgrade**: if you've upgraded R recently and see an error like `Error: package or namespace load failed for 'glmmTMB' ... unable to load shared object`, reinstalling `TMB` and `glmmTMB` from source with `--no-multiarch` (via `install.packages(type = "source")`) usually fixes it.
- **Missing compiler toolchain**: source installs of `TMB`/`glmmTMB` need a working C++ compiler. On Windows this means Rtools; on macOS, Xcode command line tools. The check below tells you if compilation is unavailable *before* you hit a cryptic error.

```{r}
#| label: install-packages
#| eval: false

required_pkgs <- c("tidyverse", "glmmTMB", "marginaleffects", "patchwork", "GGally", "distributional", "TMB", "Matrix")

# Reinstalling Matrix/TMB/glmmTMB together (rather than one at a time) avoids the
# most common cause of glmmTMB load failures: an ABI mismatch between glmmTMB's
# compiled code and the Matrix/TMB versions it was linked against.
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

# Verify glmmTMB actually loads and can fit a trivial model - this is the step
# that catches ABI mismatches that "successful" installation can still leave behind.
glmmtmb_ok <- tryCatch({
  library(glmmTMB)
  test_fit <- glmmTMB(mpg ~ wt, data = mtcars)
  TRUE
}, error = function(e) {
  message("glmmTMB failed to load/fit: ", conditionMessage(e))
  FALSE
})

if (!glmmtmb_ok) {
  message("Reinstalling TMB and glmmTMB from source to resolve a likely ABI mismatch...")
  install.packages(c("TMB", "glmmTMB"), type = "source")

  # Re-check after the source reinstall
  glmmtmb_ok <- tryCatch({
    library(glmmTMB)
    glmmTMB(mpg ~ wt, data = mtcars)
    TRUE
  }, error = function(e) {
    message("Still failing: ", conditionMessage(e))
    message("If this persists, check that you have a working C++ compiler installed ",
            "(Rtools on Windows, Xcode command line tools on macOS), then restart R and re-run this chunk.")
    FALSE
  })
}

stopifnot("glmmTMB did not install/load successfully - see messages above" = glmmtmb_ok)
```

## Setup

Run this at the start of every session (after the install chunk above has succeeded once).

```{r}
#| label: setup
library(tidyverse)       # Data manipulation and plotting
library(glmmTMB)         # Generalized linear (mixed) models
library(marginaleffects) # Compute marginal/conditional effects
library(patchwork)       # Combine multiple plots
library(GGally)          # Pairwise plots
library(distributional)  # Simulate mixture/gamma/normal distributions (Part 3)

theme_set(theme_minimal(base_size = 14))

set.seed(4127)
```

## Coding style & functions used throughout

Before diving into the case studies, here's a quick reference for the conventions and workhorse functions used repeatedly in this handout.

**The native pipe, `|>`.** Wherever you see `data |> mutate(...) |> ggplot(...)`, read it left-to-right as "take `data`, *and then* apply `mutate()`, *and then* pass the result to `ggplot()`". It's equivalent to nesting the calls (`ggplot(mutate(data, ...))`), just easier to read and edit in sequence. We use the base-R pipe (`|>`), not the older `magrittr` pipe (`%>%`) — for our purposes here they behave the same way.

**Simulating data with `tibble()` and random-draw functions.** Every case study starts by *simulating* data rather than loading a real dataset. We pick "true" parameter values (an intercept, one or more slopes) up front, then generate observations from a known random process:

- `rnorm(n, mean, sd)` — draws from a Normal distribution (continuous outcomes, e.g. weight, petal length)
- `rbinom(n, size = 1, prob)` — draws from a Bernoulli/Binomial distribution (presence/absence outcomes)
- `rpois(n, lambda)` — draws from a Poisson distribution (count outcomes, e.g. abundance)
- `runif(n, min, max)` — draws from a Uniform distribution (often used to simulate a predictor, not an outcome)

Simulating this way means we *know the ground truth* (the exact intercept and slopes used to generate the data), so we can directly check whether a fitted model recovers it — a habit worth carrying into your own work whenever you want to sanity-check a modeling approach before applying it to real data.

**Fitting models with `glmmTMB()`.** Our model-fitting workhorse throughout is `glmmTMB(response ~ predictors, data = ..., family = ...)`. Key things to know:

- `y ~ x1 * x2` expands to `y ~ x1 + x2 + x1:x2` — main effects for both predictors *and* their interaction, all in one shorthand
- `family = gaussian()` for continuous outcomes, `binomial(link = "logit")` for presence/absence, `poisson(link = "log")` for counts
- `dispformula = ~ x` lets predictors affect the *variance* of the outcome, not just its mean (used in the distributional-regression section of Part 2)

**The `marginaleffects` package.** Raw model coefficients (from `summary()`) are often not the quantity you actually want to report — especially with interactions or non-linear link functions. `marginaleffects` converts coefficients into direct, interpretable answers:

- `predictions()` — the model-implied value of the outcome at specific covariate values (a **conditional** prediction)
- `avg_predictions()` — the same, but averaged over a distribution of covariate values (a **marginal** prediction)
- `slopes()` — the model-implied *effect* (derivative/slope) of one predictor on the outcome, at specific covariate values
- `avg_slopes()` — the same, averaged over a distribution of covariate values
- `datagrid(...)` — builds the covariate grid at which `predictions()`/`slopes()` are evaluated; arguments can be fixed values, functions like `mean` or `fivenum` (the [five-number summary](https://en.wikipedia.org/wiki/Five-number_summary)), or vectors of values
- `by = "group"` — instead of collapsing (averaging) over a variable, keep results broken out by its levels
- `hypotheses(hypothesis = "b1 - b2 = 0")` — tests linear combinations of estimates, e.g. "is the slope for group A different from the slope for group B?"

**`set.seed()`.** Called once near the top of each script (or before an exercise you want to be able to compare answers on) so that "random" simulation is actually reproducible — running the same code twice gives identical numbers. If your results differ from a classmate's despite identical code, check whether you both used the same seed at the same point in the script.

# Part 1 — Categorical × Numerical Interactions

## Case study: Bighorn sheep

Ram Mountain, Alberta. Question: how does winter snow depth relate to spring mass in Bighorn sheep, and does it differ by sex? Males and females segregate in winter — females use lower, less snowy habitat.

```{r}
#| label: simulate-bighorn
n_individuals_population <- 1000
sexes <- c("male", "female")
male_intercept <- 220; female_intercept <- 180
sd_weight <- 15
male_loss <- -0.8; female_loss <- -2.5

bighorn_population <- tibble(
  sex = rep(sexes, n_individuals_population / length(sexes)),
  snow_depth = runif(n_individuals_population,
    min = ifelse(sex == "male", 35, 0),
    max = ifelse(sex == "male", 70, 35)
  ),
  weight = rnorm(n_individuals_population,
    mean = ifelse(sex == "male",
      male_intercept + male_loss * snow_depth,
      female_intercept + female_loss * snow_depth),
    sd = sd_weight
  )
)
```

Built-in features: different intercepts, different slopes, **and** different snow-depth distributions by sex.

## Simpson's Paradox

```{r}
#| label: simpsons-paradox
#| fig-cap: "Ignoring sex hides the true within-sex trends."
ggplot(bighorn_population, aes(x = snow_depth, y = weight)) +
  geom_point(aes(color = sex)) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1)
```

Pooled: snow depth looks **positively** associated with weight — even though both sexes lose weight with more snow.

```{r}
#| label: within-sex-trends
#| fig-cap: "Within-sex trends are negative, as expected."
ggplot(bighorn_population, aes(x = snow_depth, y = weight, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## 🧪 Exercise 1

Using `bighorn_population`:

1.  Fit `weight ~ sex * snow_depth` with `glmmTMB()`. Call it `mod_mass_true`.
2.  From the summary, identify which coefficient represents the **male** slope of snow depth. (Hint: it's not printed directly — you have to add two rows together.)
3.  In one sentence, explain why a naive `weight ~ snow_depth` model (ignoring sex) would mislead a manager trying to understand the effect of a harsh winter.

```{r}
#| label: exercise-1

```

## Sampling from the population

In practice we can't measure everyone. Two sampling scenarios:

- **Scenario 1**: balanced trap, 50 males / 50 females
- **Scenario 2**: biased trap, 80 males / 20 females

```{r}
#| label: sample-scenario-1
n_individuals <- 100
bighorn_mass_1 <- bighorn_population |>
  group_by(sex) |>
  slice_sample(n = n_individuals / 2, replace = FALSE) |>
  ungroup()

mod_mass_1 <- glmmTMB(weight ~ sex * snow_depth,
                       data = bighorn_mass_1, family = gaussian())
summary(mod_mass_1)$coefficients$cond
```

Reading the coefficients:

- **Intercept**: female weight at snow_depth = 0
- **sexmale**: male − female intercept difference
- **snow_depth**: female slope
- **sexmale:snow_depth**: male − female slope difference

## Q1: Do sexes differ in spring mass?

```{r}
#| label: q1-sex-difference
predictions(mod_mass_1, by = "sex")

predictions(mod_mass_1, by = "sex") |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

## Q2: Does snow depth affect mass equally by sex?

```{r}
#| label: q2-slopes-by-sex
#| fig-cap: "Slopes of snow depth by sex."
slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  ggplot(aes(x = estimate, xmax = conf.high, xmin = conf.low, y = sex, color = sex)) +
  geom_pointrange(size = 1) +
  theme(legend.position = "none")

slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

## Q3: What's the population-average slope?

`avg_slopes()` computes the slope for every observation, then averages:

```{r}
#| label: q3-avg-slope
avg_slopes(mod_mass_1, variables = "snow_depth", vcov = TRUE)
```

## Scenario 2: biased sampling (80% male)

```{r}
#| label: sample-scenario-2
bighorn_mass_2 <- bind_rows(
  bighorn_population |> filter(sex == "male") |> slice_sample(n = 80),
  bighorn_population |> filter(sex == "female") |> slice_sample(n = 20)
)
mod_mass_2 <- glmmTMB(weight ~ sex * snow_depth, data = bighorn_mass_2, family = gaussian())
avg_slopes(mod_mass_2, variables = "snow_depth", vcov = TRUE)
```

Conditional (by-sex) slopes stay close to truth. But the **marginal** slope is biased toward the overrepresented, weaker-effect sex (males).

## Correcting for known sampling bias

If we *know* the true population sex ratio (50:50), we can average over a **balanced** `newdata` grid instead of our biased sample.

```{r}
#| label: balanced-newdata
newdata_balanced <- bighorn_mass_2 |>
  group_by(sex) |>
  reframe(snow_depth = seq(min(snow_depth), max(snow_depth), length.out = 50)) |>
  ungroup()

avg_predictions(mod_mass_2, variables = "snow_depth", newdata = newdata_balanced) |>
  as_tibble() |> select(estimate, conf.low, conf.high) |> head(3)
```

## 🧪 Exercise 2

Using `mod_mass_2` (the biased sample model):

1.  Compute `avg_slopes()` **without** correcting for sampling bias, and again **with** the balanced `newdata_balanced` grid.
2.  Which one is closer to the true average slope, `mean(c(male_loss, female_loss))`?
3.  Suppose instead you *didn't* know the true sex ratio in the population — what additional data would you need to collect to justify a correction like this?

```{r}
#| label: exercise-2

```

# Part 2 — Numerical × Numerical Interactions

## Case study: petal length

Simulated flowers; petal length as a function of temperature and rain, with an interaction. Predictors are **mean-centered**: the intercept becomes the predicted response at *average* conditions rather than at (biologically meaningless) zero temperature/rain.

```{r}
#| label: simulate-flowers
n_flowers <- 200
temperature <- rnorm(n_flowers, mean = 10, sd = 15)
rain <- rgamma(n_flowers, shape = 2)
avg_petal_length <- 25.6; sd_petal_length <- 4.5
temperature_effect <- 0.4; rain_effect <- 0.9; interaction_effect <- 0.2

flower_measurements_1 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered +
      interaction_effect * temp_centered * rain_centered,
    sd = sd_petal_length)
)

mod_flower_1 <- glmmTMB(petal_length ~ temp_centered * rain_centered,
                         data = flower_measurements_1)
summary(mod_flower_1)$coefficients$cond
```

## Slope of one predictor at a single value of the other

```{r}
#| label: slope-at-average-rain
# Effect of temperature when rain_centered = 0 (i.e., average rain)
slopes(mod_flower_1, newdata = datagrid(rain_centered = 0),
       variables = "temp_centered")
```

This matches the `temp_centered` coefficient directly — because rain is centered, "at rain = 0" means "at average rain."

## Slope across the range of the moderator

```{r}
#| label: slope-across-temperature
#| fig-cap: "Effect of rain on petal length, across observed temperatures."
slopes_by_temp <- slopes(mod_flower_1,
  newdata = datagrid(temp_centered = seq(min(flower_measurements_1$temp_centered),
                                          max(flower_measurements_1$temp_centered), length.out = 30)),
  variables = "rain_centered", by = "temp_centered")

slopes_by_temp |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linetype = 2) +
  geom_ribbon(aes(x = temp_centered + mean(temperature), ymin = conf.low, ymax = conf.high),
    alpha = 0.4, fill = "lightgrey") +
  geom_line(aes(x = temp_centered + mean(temperature), y = estimate), linewidth = 1) +
  labs(y = "Effect of rain on petal length", x = "Temperature values")
```

The interaction coefficient is exactly the slope of this line — how much the effect of rain *changes* per unit of temperature.

## Visualizing predictions at chosen moderator values

Common choice: the [five-number summary](https://en.wikipedia.org/wiki/Five-number_summary) of the moderator.

```{r}
#| label: five-number-summary-predictions
#| fig-cap: "Predicted petal length across temperature, at five reference rain levels."
predictions(mod_flower_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
                         max(flower_measurements_1$temp_centered), length.out = 30),
    rain_centered = fivenum
  )) |>
  as_tibble() |>
  mutate(rain = round(rain_centered + mean(rain), 2)) |>
  ggplot() +
  geom_line(aes(y = estimate, x = temp_centered + mean(temperature), color = as.factor(rain))) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high,
                  x = temp_centered + mean(temperature), fill = as.factor(rain)), alpha = 0.2) +
  labs(y = "Predicted petal length", x = "Temperature", color = "Rain", fill = "Rain")
```

## 🧪 Exercise 3

Using `mod_flower_1`:

1.  Compute the slope of **temperature** across the observed range of **rain** (mirror the rain-across-temperature plot above).
2.  At what value of rain does the temperature effect become non-significant (CI crosses zero), if at all?
3.  Rewrite the research question this model could answer as a single sentence, identifying which variable is the *exposure* and which is the *moderator*.

```{r}
#| label: exercise-3

```

## Interactions beyond the mean: distributional regression

Interactions don't have to act only on the average response. `glmmTMB`'s `dispformula` lets predictors affect the **standard deviation** too — useful when variance itself is ecologically meaningful (e.g., reduced trait variance under stress).

```{r}
#| label: simulate-heteroscedastic
log_sd_intercept <- log(sd_petal_length)
log_sd_temp_slope <- 0.03

flower_measurements_2 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length + temperature_effect * temp_centered,
    sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))
)

mod_flower_2 <- glmmTMB(petal_length ~ temp_centered,
                         dispformula = ~ temp_centered,
                         data = flower_measurements_2)
```

```{r}
#| label: check-dispersion-model
#| fig-cap: "Recovered vs. simulated standard deviation of petal length."
pred_sd_temp <- predictions(mod_flower_2, newdata = flower_measurements_2, type = "disp") |>
  as_tibble()

true_sd_temp <- flower_measurements_2 |>
  mutate(true_sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))

ggplot() +
  geom_line(data = true_sd_temp,
            aes(x = temp_centered + mean(temperature), y = true_sd, color = "Simulated"),
            linewidth = 1, linetype = "dashed") +
  geom_ribbon(data = pred_sd_temp,
              aes(ymin = conf.low, ymax = conf.high, x = temp_centered + mean(temperature)),
              fill = "red", alpha = 0.3) +
  geom_line(data = pred_sd_temp,
            aes(x = temp_centered + mean(temperature), y = estimate, color = "Estimated"),
            linewidth = 1) +
  scale_color_manual(values = c("Simulated" = "darkgreen", "Estimated" = "red"), name = "Standard deviation") +
  labs(x = "Temperature", y = "Standard deviation")
```

`type = "disp"` retrieves predictions/slopes for the dispersion sub-model, exactly as `type = "conditional"` does for the mean.

## 🧪 Exercise 4 (challenge)

Extend `mod_flower_2` by adding `rain_centered` and a `temp_centered * rain_centered` interaction to **both** the conditional formula and the `dispformula`.

1.  Fit the model on simulated data of your own design (or reuse the logic from `flower_measurements_2`, adding a rain effect and interaction term on the log-sd scale).
2.  Produce one plot showing how the effect of rain on the **standard deviation** of petal length changes across temperature.
3.  In plain language, describe a real ecological scenario (not necessarily plants) where an interaction on variance, rather than the mean, would be the interesting finding.

```{r}
#| label: exercise-4

```

# Part 3 — Link Functions Create Implicit Interactions

Non-linear link functions (logit, log, ...) introduce a *second*, implicit source of interaction: even a model with **no interaction term at all** can show predictors whose response-scale effect depends on the values of other predictors. This happens because the link function bends a straight line (on the *link* scale) into a curve (on the *response* scale).

## Case study: species occurrence

Simulated presence/absence of a species across 200 plots, as a function of tree cover, rain, and temperature — **no interaction terms**, fit with a binomial GLM (logit link).

```{r}
#| label: simulate-occurrence
n_plots <- 200
tree_cover <- dist_mixture(dist_beta(4, 0.8), dist_beta(0.8, 4), weights = c(0.5, 0.5))
avg_yearly_rain <- dist_gamma(shape = 3, rate = 0.7)
avg_yearly_temp <- dist_normal(mu = 20, sd = 10)

occurence_intercept <- plogis(-3)
occurence_tree_effect <- 0.45
occurence_rain_effect <- 0.01
occurence_temp_effect <- 0.01

occ_data <- tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |>
  unnest(everything()) |>
  mutate(occurence = rbinom(n_plots, size = 1,
    prob = occurence_intercept +
      occurence_tree_effect * tree_cov +
      occurence_rain_effect * rain +
      occurence_temp_effect * temp))
```

```{r}
#| label: fit-occurrence-model
mod_1 <- glmmTMB(occurence ~ tree_cov + rain + temp, family = binomial(link = "logit"), data = occ_data)
summary(mod_1)$coefficients$cond
```

We simulated the effect of tree cover to be **0.45** on the probability scale, but the model reports **\~1.87**. That's not an error — it's reporting the coefficient on the **log-odds** scale, not the probability scale.

## The logit link bends a line into an S-shape

```{r}
#| label: logistic-curve
#| fig-cap: "The logit link bends a straight line into an S-shape."
tibble(log_odds = seq(-6, 6, length.out = 400)) |>
  mutate(probability = plogis(log_odds)) |>
  ggplot(aes(x = log_odds, y = probability)) +
  geom_line(linewidth = 1) +
  labs(x = "log-odds scale (what the coefficient describes)",
       y = "probability scale (what we care about)")
```

Near the middle (p ≈ 0.5) the curve is steep: a small log-odds push moves probability a lot. Near the ends (p ≈ 0 or 1) the curve flattens: the same push barely moves probability.

## From log-odds to probability, by hand

The derivative of the inverse-logit gives the conversion: if $\eta$ is the log-odds and $p = \text{logit}^{-1}(\eta)$,

$$\frac{dp}{dX} = p(1-p) \times \beta$$

```{r}
#| label: hand-derive-probability-slope
beta_tree <- fixef(mod_1)$cond["tree_cov"]
eta <- predict(mod_1, newdata = datagrid(model = mod_1, temp = mean(occ_data$temp),
                                          rain = mean(occ_data$rain)), type = "link")
p <- plogis(eta)
p * (1 - p) * beta_tree  # compare to slopes() below, and to our simulated 0.45
```

`marginaleffects` does this same conversion for us directly:

```{r}
#| label: slopes-probability-scale
slopes(mod_1, variables = "tree_cov",
       newdata = datagrid(temp = mean(occ_data$temp), rain = mean(occ_data$rain)))
```

## The implicit interaction, visualized

```{r}
#| label: implicit-interaction-plot
#| fig-cap: "The tree-cover slope on the probability scale changes across rain x temp, though no interaction term exists in the model."
slopes(mod_1, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) |>
  ggplot() +
  geom_pointrange(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = 1)) +
  geom_vline(xintercept = occurence_tree_effect, alpha = .8) +
  facet_grid(rain ~ temp) +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank())
```

Because the amount of "bending" depends on where you are on the S-curve, tree cover's probability-scale effect depends on rain and temp too — even though we never told the model to include that interaction. Whether an interaction "exists" can depend on which scale you look at (additive/probability vs. multiplicative/log-odds) — see Spake et al. (2023) for more on this "Type-D" error.

## 🧪 Exercise 5

Using `mod_1`:

1.  Compute the probability-scale slope of `tree_cov` at the *minimum* and *maximum* observed values of `rain` and `temp` (four combinations total, via `datagrid()`).
2.  Which combination gives the smallest slope? Which gives the largest? Relate your answer to where those combinations fall on the S-curve.
3.  In one sentence, explain why reporting only the raw `summary(mod_1)` coefficient for `tree_cov` would be misleading to a reader who cares about probability of occurrence.

```{r}
#| label: exercise-5

```

## Case study: species abundance

The same phenomenon shows up with **count** data. Abundance (when present) as a function of the same three predictors, fit with a Poisson GLM (log link), again **no interaction terms**.

```{r}
#| label: simulate-abundance
abundance_intercept <- exp(1)
abundance_tree_effect <- 4.5
abundance_rain_effect <- 0.1
abundance_temp_effect <- 0.1

abund_data <- tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |>
  unnest(everything()) |>
  mutate(abundance = rpois(n_plots, lambda =
    abundance_intercept +
    abundance_tree_effect * tree_cov +
    abundance_rain_effect * rain +
    abundance_temp_effect * temp))

mod_2 <- glmmTMB(abundance ~ tree_cov + rain + temp, family = poisson(link = "log"), data = abund_data)
summary(mod_2)$coefficients$cond
```

The log-scale `tree_cov` coefficient is **\~0.73**, not 4.5.

## The log link bends a line into an exponential curve

```{r}
#| label: exponential-curve
#| fig-cap: "The log link bends a straight line into an exponential curve."
tibble(log_count = seq(-2, 6, length.out = 400)) |>
  mutate(count = exp(log_count)) |>
  ggplot(aes(x = log_count, y = count)) +
  geom_line(linewidth = 1) +
  labs(x = "log scale (what the coefficient describes)", y = "count scale (what we care about)")
```

The local rate of change on the count scale is $$\lambda \cdot \beta$$ (the expected count itself, times the log-scale coefficient) — the count-scale analogue of $$p(1-p)\cdot\beta$$.

## Two ways to report a log-link coefficient

- **Rate ratio** ($$e^\beta$$): a *constant* multiplicative effect, valid everywhere — "each unit increase in tree cover multiplies expected abundance by this factor, regardless of rain or temp"
- **Local slope** ($$\lambda \cdot \beta$$): an *additive*, count-scale effect ("this many more individuals"), but only valid near the specific covariate values used

```{r}
#| label: rate-ratio-vs-local-slope
exp(fixef(mod_2)$cond["tree_cov"])  # rate ratio: ~2.08, i.e. a ~108% increase per unit tree cover

beta_tree_2 <- fixef(mod_2)$cond["tree_cov"]
eta_2 <- predict(mod_2, newdata = datagrid(model = mod_2, temp = mean(abund_data$temp),
                                            rain = mean(abund_data$rain)), type = "link")
lambda <- exp(eta_2)
lambda * beta_tree_2  # local count-scale slope, close to our simulated 4.5
```

Both describe the same underlying model — pick the one that matches your audience (percent change vs. a concrete headcount).

## 🧪 Exercise 6

Using `mod_2`:

1.  Compute the rate ratio and the local count-scale slope for `temp`, holding `rain` at its minimum (a "drought" scenario) and `tree_cov` at its mean.
2.  Repeat holding `tree_cov` at a low vs. a high value (e.g., the 10th and 90th percentiles) — does the rate ratio change? Does the local slope?
3.  Explain, in your own words, why one of these two summaries stays constant across `tree_cov` and the other doesn't.

```{r}
#| label: exercise-6

```

## Further reading

- Full write-up with additional detail: [Part 3 tutorial](../../posts/interactions_link_functions/index.qmd)
- `marginaleffects` documentation: <https://marginaleffects.com/>
- Spake et al. (2023), on interactions in GLMs: <https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12939>