---
title: "Interactions in Regression Models — Student Handout"
subtitle: "Follow-along script and exercises"
author: "Your name here"
date: 10-05-2026
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: false
---

## Before you start

Run the chunk below **once** to install everything you need. It's written to avoid the most common `glmmTMB` installation failures:

- **Binary vs. source mismatch**: `glmmTMB` links against `TMB` and `Matrix`, which must be built with compatible ABI versions. Installing all three together from the same repository snapshot (rather than upgrading one in isolation) avoids "package was installed with a different version" load-time errors.
- **Stale compiled objects after an R or `Matrix` upgrade**: if you've upgraded R recently and see an error like `Error: package or namespace load failed for 'glmmTMB' ... unable to load shared object`, reinstalling `TMB` and `glmmTMB` from source with `--no-multiarch` (via `install.packages(type = "source")`) usually fixes it.
- **Missing compiler toolchain**: source installs of `TMB`/`glmmTMB` need a working C++ compiler. On Windows this means Rtools; on macOS, Xcode command line tools. The check below tells you if compilation is unavailable *before* you hit a cryptic error.

```{r}
#| label: install-packages
#| eval: false

required_pkgs <- c("tidyverse", "glmmTMB", "marginaleffects", "patchwork", "GGally",
                    "distributional", "ggdist", "tinytable", "performance", "TMB", "Matrix")

# Reinstalling Matrix/TMB/glmmTMB together (rather than one at a time) avoids the
# most common cause of glmmTMB load failures: an ABI mismatch between glmmTMB's
# compiled code and the Matrix/TMB versions it was linked against.
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

# Verify glmmTMB actually loads and can fit a trivial model - this is the step
# that catches ABI mismatches that "successful" installation can still leave behind.
glmmtmb_ok <- tryCatch({
  library(glmmTMB)
  test_fit <- glmmTMB(mpg ~ wt, data = mtcars)
  TRUE
}, error = function(e) {
  message("glmmTMB failed to load/fit: ", conditionMessage(e))
  FALSE
})

if (!glmmtmb_ok) {
  message("Reinstalling TMB and glmmTMB from source to resolve a likely ABI mismatch...")
  install.packages(c("TMB", "glmmTMB"), type = "source")

  # Re-check after the source reinstall
  glmmtmb_ok <- tryCatch({
    library(glmmTMB)
    glmmTMB(mpg ~ wt, data = mtcars)
    TRUE
  }, error = function(e) {
    message("Still failing: ", conditionMessage(e))
    message("If this persists, check that you have a working C++ compiler installed ",
            "(Rtools on Windows, Xcode command line tools on macOS), then restart R and re-run this chunk.")
    FALSE
  })
}

stopifnot("glmmTMB did not install/load successfully - see messages above" = glmmtmb_ok)
```

## Setup

Run this at the start of every session (after the install chunk above has succeeded once).

```{r}
#| label: setup
library(tidyverse)       # Data manipulation and plotting
library(glmmTMB)         # Generalized linear (mixed) models
library(marginaleffects) # Compute marginal/conditional effects
library(patchwork)       # Combine multiple plots
library(GGally)          # Pairwise plots
library(distributional)  # Simulate mixture/gamma/normal distributions (Part 3)
library(ggdist)          # stat_slab() for predicted-distribution plots (Part 2)
library(tinytable)       # tt(): simple, nice-looking results tables
library(performance)     # r2(), r2_tjur(): model R-squared measures

theme_set(theme_minimal(base_size = 14))

# silence marginaleffects warnings about glmmTMB
options(marginaleffects_safe = FALSE)

set.seed(333)
```

## Coding style & functions used throughout

Before diving into the case studies, here's a quick reference for the conventions and workhorse functions used repeatedly in this handout.

**The native pipe, `|>`.** Wherever you see `data |> mutate(...) |> ggplot(...)`, read it left-to-right as "take `data`, *and then* apply `mutate()`, *and then* pass the result to `ggplot()`". It's equivalent to nesting the calls (`ggplot(mutate(data, ...))`), just easier to read and edit in sequence. We use the base-R pipe (`|>`), not the older `magrittr` pipe (`%>%`) — for our purposes here they behave the same way.

**Simulating data with `tibble()` and random-draw functions.** Every case study starts by *simulating* data rather than loading a real dataset. We pick "true" parameter values (an intercept, one or more slopes) up front, then generate observations from a known random process:

- `rnorm(n, mean, sd)` — draws from a Normal distribution (continuous outcomes, e.g. weight, petal length)
- `rbinom(n, size = 1, prob)` — draws from a Bernoulli/Binomial distribution (presence/absence outcomes)
- `rpois(n, lambda)` — draws from a Poisson distribution (count outcomes, e.g. abundance)
- `runif(n, min, max)` — draws from a Uniform distribution (often used to simulate a predictor, not an outcome)
- `distributional::dist_*()` + `generate()` — an alternative way to simulate predictors (e.g. `dist_beta()`, `dist_gamma()`, `dist_mixture()`) used in Part 3

Simulating this way means we *know the ground truth* (the exact intercept and slopes used to generate the data), so we can directly check whether a fitted model recovers it — a habit worth carrying into your own work whenever you want to sanity-check a modeling approach before applying it to real data.

**Fitting models with `glmmTMB()`.** Our model-fitting workhorse throughout is `glmmTMB(response ~ predictors, data = ..., family = ...)`. Key things to know:

- `y ~ x1 * x2` expands to `y ~ x1 + x2 + x1:x2` — main effects for both predictors *and* their interaction, all in one shorthand
- `family = gaussian()` for continuous outcomes, `binomial(link = "logit")` for presence/absence, `poisson(link = "log")` for counts, `truncated_poisson` for a hurdle model's count-when-present part
- `dispformula = ~ x` lets predictors affect the *variance* of the outcome, not just its mean (used in the distributional-regression section of Part 2)
- `zi = ~ x` adds a zero-inflation sub-model: a binomial part that models whether a value is a "structural zero" (used in the hurdle-model section of Part 3)

**The `marginaleffects` package.** Raw model coefficients (from `summary()`) are often not the quantity you actually want to report — especially with interactions or non-linear link functions. `marginaleffects` converts coefficients into direct, interpretable answers:

- `predictions()` — the model-implied value of the outcome at specific covariate values (a **conditional** prediction)
- `avg_predictions()` — the same, but averaged over a distribution of covariate values (a **marginal** prediction)
- `slopes()` — the model-implied *effect* (derivative/slope) of one predictor on the outcome, at specific covariate values
- `avg_slopes()` — the same, averaged over a distribution of covariate values
- `datagrid(...)` — builds the covariate grid at which `predictions()`/`slopes()` are evaluated; arguments can be fixed values, functions like `mean` or `fivenum` (the [five-number summary](https://en.wikipedia.org/wiki/Five-number_summary)), or vectors of values
- `datagrid(..., grid_type = "counterfactual")` — pairs each focal value with *every observed row* of the data, so `avg_predictions()`/`avg_slopes()` genuinely average over the observed distribution of the other predictors, instead of holding them fixed at a single value (e.g., their mean)
- `by = "group"` — instead of collapsing (averaging) over a variable, keep results broken out by its levels
- `type = "response"` vs. `type = "link"` — get a predictor's effect on the natural (probability, count) scale vs. the scale the model is additive on; for hurdle/zero-inflated models, `type = "conditional"`, `"response"`, and `"zprob"` select which model component to report
- `hypotheses(hypothesis = "b1 - b2 = 0")` — tests linear combinations of estimates, e.g. "is the slope for group A different from the slope for group B?"

**`set.seed()`.** Called once near the top of this script so that "random" simulation is actually reproducible — running the same code twice gives identical numbers. If your results differ from a classmate's despite identical code, check whether you both used the same seed at the same point in the script.

# Part 1 — Categorical × Numerical Interactions

## What is an interaction?

An interaction means **the effect of one predictor depends on the value of another**.

- No interaction: snow depth affects weight the same way for every sheep
- Interaction: the effect of snow depth on weight *differs by sex*

For a categorical moderator (e.g., sex) and a continuous predictor (e.g., snow depth): the interaction means the **slope** differs between categories.

## Conditional vs. marginal effects

- **Conditional effects**: the effect of a predictor *at specific values* of other predictors (e.g., the slope of snow depth *for males*)
- **Marginal effects**: the effect *averaged across* the distribution of other predictors (e.g., the *average* slope of snow depth across the whole population)

Both are legitimate; they answer different questions. Interactions are exactly where the distinction starts to matter.

## Case study: Bighorn sheep

Ram Mountain, Alberta. Question: how does winter snow depth relate to spring mass in Bighorn sheep, and does it differ by sex? Males and females segregate in winter — females use lower, less snowy habitat.

```{r}
#| label: simulate-bighorn
n_individuals_population <- 1000
sexes <- c("male", "female")
male_intercept <- 220; female_intercept <- 180
sd_weight <- 15
male_loss <- -0.8; female_loss <- -2.5

bighorn_population <- tibble(
  sex = rep(sexes, n_individuals_population / length(sexes)),
  snow_depth = runif(n_individuals_population,
    min = ifelse(sex == "male", 35, 0),
    max = ifelse(sex == "male", 70, 35)
  ),
  weight = rnorm(n_individuals_population,
    mean = ifelse(sex == "male",
      male_intercept + male_loss * snow_depth,
      female_intercept + female_loss * snow_depth),
    sd = sd_weight
  )
)
```

Built-in features: different intercepts, different slopes, **and** different snow-depth distributions by sex.

## Simpson's Paradox

Pooling males and females together and fitting a single trend line can hide — or even reverse — the true relationship.

```{r}
#| label: simpsons-paradox
#| fig-cap: "Ignoring sex hides the true within-sex trends."
ggplot(bighorn_population, aes(x = snow_depth, y = weight)) +
  geom_point(aes(color = sex)) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1)
```

Pooled: snow depth looks **positively** associated with weight — even though both sexes lose weight with more snow.

## Why? Sex is a confounder

- Within each sex: more snow → lower weight (true negative relationship)
- Between sexes: males see more snow **and** are heavier (confounded)
- Pooled data compares heavy, high-snow males to light, low-snow females → spurious positive trend

```{r}
#| label: within-sex-trends
#| fig-cap: "Within-sex trends are negative, as expected."
ggplot(bighorn_population) +
  geom_point(aes(x = snow_depth, y = weight, color = sex)) +
  geom_smooth(aes(x = snow_depth, y = weight), method = "lm", se = FALSE, color = "red") +
  geom_smooth(aes(x = snow_depth, y = weight, color = sex), method = "lm", se = FALSE)
```

## Exercise 1

Using `bighorn_population`:

1.  Fit `weight ~ sex * snow_depth` with `glmmTMB()`. Call it `mod_mass_true`.
2.  From the summary, identify which coefficient represents the **male** slope of snow depth. (Hint: it's not printed directly — you have to add two rows together.)
3.  In one sentence, explain why a naive `weight ~ snow_depth` model (ignoring sex) would mislead a manager trying to understand the effect of a harsh winter. Can you think of an example of such a situation in your own field or research?

*(\~5 minutes — discuss with a neighbor)*

```{r}
#| label: exercise-1

```

## Sampling from the population

In practice we can't measure everyone. Two sampling scenarios:

- **Scenario 1**: balanced trap, 50 males / 50 females
- **Scenario 2**: biased trap, 80 males / 20 females

```{r}
#| label: sample-scenario-1
n_individuals <- 100
bighorn_mass_1 <- bighorn_population |>
  group_by(sex) |>
  slice_sample(n = n_individuals / 2, replace = FALSE) |>
  ungroup()

mod_mass_1 <- glmmTMB(weight ~ sex * snow_depth,
                       data = bighorn_mass_1, family = gaussian())
summary(mod_mass_1)$coefficients$cond
```

Reading the coefficients:

- **Intercept**: female weight at snow_depth = 0
- **sexmale**: male − female intercept difference
- **snow_depth**: female slope
- **sexmale:snow_depth**: male − female slope difference

Useful for hypothesis tests, not directly interpretable for "what's the effect for males?" — that's what `marginaleffects` is for.

## Q1: Do sexes differ in spring mass?

```{r}
#| label: q1-sex-difference
predictions(mod_mass_1, by = "sex")

predictions(mod_mass_1, by = "sex") |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

This is a **marginal prediction**: the model-implied mean for each sex, averaged over the observed snow-depth distribution in the sample.

## Q2: Does snow depth affect mass equally by sex?

```{r}
#| label: q2-slopes-by-sex
#| fig-cap: "Slopes of snow depth by sex, with true simulated values as vertical lines."
slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  ggplot(aes(x = estimate, xmax = conf.high, xmin = conf.low, y = sex, color = sex)) +
  geom_pointrange(size = 1) +
  geom_vline(xintercept = female_loss, color = "red") +
  geom_vline(xintercept = male_loss, color = "blue") +
  theme(legend.position = "none")

slopes(mod_mass_1, variables = "snow_depth", by = "sex", vcov = TRUE) |>
  hypotheses(hypothesis = "b1 - b2 = 0")
```

## Q3: What's the population-average slope?

`avg_slopes()` computes the slope for every observation, then averages:

```{r}
#| label: q3-avg-slope
avg_slopes(mod_mass_1, variables = "snow_depth", vcov = TRUE)

mean(c(male_loss, female_loss))  # true simulated average
```

## Marginal vs conditional, visualized

```{r}
#| label: marginal-vs-conditional
#| fig-cap: "Green = marginal; red/blue = conditional (by sex)."
avg_pred_mass_1 <- avg_predictions(mod_mass_1, variables = "snow_depth")
pred_sex_mass_1 <- predictions(mod_mass_1, newdata = bighorn_mass_1)

ggplot() +
  geom_ribbon(data = avg_pred_mass_1,
              aes(x = snow_depth, ymin = conf.low, ymax = conf.high),
              color = "lightgreen", fill = "lightgreen", alpha = 0.4) +
  geom_line(data = avg_pred_mass_1, aes(x = snow_depth, y = estimate),
            color = "lightgreen") +
  geom_ribbon(data = pred_sex_mass_1,
              aes(x = snow_depth, ymin = conf.low, ymax = conf.high, color = sex, fill = sex),
              alpha = 0.8) +
  geom_line(data = pred_sex_mass_1, aes(x = snow_depth, y = estimate, color = sex)) +
  geom_point(data = bighorn_mass_1, aes(y = weight, x = snow_depth, color = sex)) +
  guides(alpha = "none", fill = "none") +
  labs(title = "Green = marginal; red/blue = conditional (by sex)") +
  coord_cartesian(ylim = c(50, 220))
```

## Scenario 2: biased sampling (80% male)

```{r}
#| label: sample-scenario-2
bighorn_mass_2 <- bind_rows(
  bighorn_population |> filter(sex == "male") |> slice_sample(n = 80),
  bighorn_population |> filter(sex == "female") |> slice_sample(n = 20)
)
mod_mass_2 <- glmmTMB(weight ~ sex * snow_depth, data = bighorn_mass_2, family = gaussian())
summary(mod_mass_2)
avg_slopes(mod_mass_2, variables = "snow_depth", vcov = TRUE)
```

Conditional (by-sex) slopes stay close to truth. But the **marginal** slope is biased toward the overrepresented, weaker-effect sex (males).

## Correcting for known sampling bias

If we *know* the true population sex ratio (50:50), we can average over a **balanced** `newdata` grid instead of our biased sample.

```{r}
#| label: balanced-newdata
newdata_balanced <- bighorn_mass_2 |>
  group_by(sex) |>
  reframe(snow_depth = seq(min(snow_depth), max(snow_depth), length.out = 50)) |>
  ungroup()
```

## Exercise 2

Using the model that was fit to the entire population and the one with a biased sample, use the `avg_predictions()` function along with a balanced and an unbalanced prediction grid to create a plot comparing the marginal predictions between the true population, biased population predictions, and corrected population predictions. There should be three trend lines.

*(\~15 minutes)*

```{r}
#| label: exercise-2

```

## Part 1 takeaways

1.  Interactions → conditional effects that depend on other variables
2.  Simpson's Paradox is a real risk when pooling across confounded groups
3.  Marginal ≠ conditional; both are valid but answer different questions
4.  Raw model coefficients rarely equal your research quantity of interest — use `marginaleffects`
5.  Marginal effects inherit your **sample's** covariate distribution unless you override it with `newdata`

# Part 2 — Numerical × Numerical Interactions

## Two numerical predictors

With two continuous predictors, an interaction means: **the slope of one predictor changes as the other predictor changes** — not a separate slope per group, but a sliding one.

A raw coefficient is only "the slope when the other predictor equals zero" — meaningful only if zero is a meaningful value. This motivates **centering** predictors on their mean, so the intercept and main-effect coefficients describe *average* conditions.

## Case study: petal length

Simulated flowers; petal length as a function of temperature and rain, with an interaction. Predictors are **mean-centered**: the intercept becomes the predicted response at *average* conditions rather than at (biologically meaningless) zero temperature/rain.

```{r}
#| label: simulate-flowers
n_flowers <- 200
temperature <- rnorm(n_flowers, mean = 10, sd = 15)
rain <- rgamma(n_flowers, shape = 2)
avg_petal_length <- 25.6; sd_petal_length <- 4.5
temperature_effect <- 0.4; rain_effect <- 0.9; interaction_effect <- 0.2

flower_measurements_1 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered +
      interaction_effect * temp_centered * rain_centered,
    sd = sd_petal_length)
)

mod_flower_1 <- glmmTMB(petal_length ~ temp_centered * rain_centered,
                         data = flower_measurements_1)
summary(mod_flower_1)$coefficients$cond
```

Every term has a small p-value — but what do these numbers *mean* for our research question? Time to unpack them.

## Slope of one predictor at a single value of the other

```{r}
#| label: slope-at-average-rain
# Effect of temperature when rain_centered = 0 (i.e., average rain)
slopes(mod_flower_1, newdata = datagrid(rain_centered = 0),
       variables = "temp_centered")
```

This matches the `temp_centered` coefficient directly — because rain is centered, "at rain = 0" means "at average rain."

## Slope across the *range* of the moderator

```{r}
#| label: slope-across-temperature
#| fig-cap: "Effect of rain on petal length, across observed temperatures."
slopes_by_temp <- slopes(mod_flower_1,
  newdata = datagrid(temp_centered = seq(min(flower_measurements_1$temp_centered),
                                          max(flower_measurements_1$temp_centered), length.out = 30)),
  variables = "rain_centered", by = "temp_centered")

trend_slope_temp <- coef(lm(estimate ~ temp_centered, data = slopes_by_temp))[["temp_centered"]]

slopes_by_temp |>
  ggplot() +
  geom_hline(yintercept = 0, color = "blue", linetype = 2) +
  geom_ribbon(aes(x = temp_centered + mean(temperature), ymin = conf.low, ymax = conf.high),
    alpha = 0.4, fill = "lightgrey") +
  geom_line(aes(x = temp_centered + mean(temperature), y = estimate)) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
    label = paste0("slope = ", round(trend_slope_temp, 3))) +
  labs(y = "Effect of rain on petal length", x = "Temperature values")
```

The interaction coefficient is exactly the slope of this line — how much the effect of rain *changes* per unit of temperature.

## Visualizing predictions at chosen moderator values

Common choice: the [five-number summary](https://en.wikipedia.org/wiki/Five-number_summary) of the moderator.

```{r}
#| label: five-number-summary-predictions
#| fig-cap: "Predicted petal length across temperature, at five reference rain levels."
predictions(mod_flower_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
                         max(flower_measurements_1$temp_centered), length.out = 30),
    rain_centered = fivenum
  )) |>
  as_tibble() |>
  mutate(rain = round(rain_centered + mean(rain), 2)) |>
  ggplot() +
  geom_line(aes(y = estimate, x = temp_centered + mean(temperature), color = as.factor(rain))) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high,
                  x = temp_centered + mean(temperature), fill = as.factor(rain)), alpha = 0.2) +
  labs(y = "Predicted petal length", x = "Temperature", color = "Rain", fill = "Rain")
```

## Response surfaces — use with caution

```{r}
#| label: response-surface
predictions(mod_flower_1,
  newdata = datagrid(
    temp_centered = seq(min(flower_measurements_1$temp_centered),
                         max(flower_measurements_1$temp_centered), length.out = 30),
    rain_centered = seq(min(flower_measurements_1$rain_centered),
                         max(flower_measurements_1$rain_centered), length.out = 30)
  )) |>
  as_tibble() |>
  mutate(rain = rain_centered + mean(rain), temp = temp_centered + mean(temperature)) |>
  ggplot(aes(x = rain, y = temp, z = estimate)) +
  stat_contour_filled() +
  labs(x = "Rain", y = "Temperature", fill = "Predicted petal length")
```

Contour bins are arbitrary and hide uncertainty — good for exploration, risky for publication-quality inference. Prefer line-and-ribbon plots when you can.

## Exercise 3

Using `mod_flower_1`:

1.  Compute the slope of **temperature** across the observed range of **rain** (mirror the rain-across-temperature plot above).
2.  At what value of rain does the temperature effect become non-significant (CI crosses zero), if at all?
3.  Rewrite the research question this model could answer as a single sentence, identifying which variable is the *exposure* and which is the *moderator*.

*(\~10 minutes)*

```{r}
#| label: exercise-3

```

## Interactions beyond the mean: distributional regression

Interactions don't have to act only on the average response. `glmmTMB`'s `dispformula` lets predictors affect the **standard deviation** too — useful when variance itself is ecologically meaningful (e.g., reduced trait variance under stress).

```{r}
#| label: simulate-heteroscedastic
log_sd_intercept <- log(sd_petal_length)
log_sd_temp_slope <- 0.03

flower_measurements_2 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(n_flowers,
    mean = avg_petal_length + temperature_effect * temp_centered,
    sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))
)
```

```{r}
#| label: spread-increases-with-temp
#| fig-cap: "The spread around the trend increases with temperature."
ggplot(flower_measurements_2, aes(x = temp_centered + mean(temperature), y = petal_length)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  geom_point() +
  labs(x = "Temperature", y = "Observed petal length")
```

```{r}
#| label: fit-dispersion-model
mod_flower_2 <- glmmTMB(petal_length ~ temp_centered,
                         dispformula = ~ temp_centered,
                         data = flower_measurements_2)
```

## Checking the dispersion model

```{r}
#| label: check-dispersion-model
#| fig-cap: "Recovered vs. simulated standard deviation of petal length."
pred_sd_temp <- predictions(mod_flower_2, newdata = flower_measurements_2, type = "disp") |>
  as_tibble()

true_sd_temp <- flower_measurements_2 |>
  mutate(true_sd = exp(log_sd_intercept + log_sd_temp_slope * temp_centered))

ggplot() +
  geom_line(data = true_sd_temp,
            aes(x = temp_centered + mean(temperature), y = true_sd, color = "Simulated"),
            linetype = "dashed") +
  geom_ribbon(data = pred_sd_temp,
              aes(ymin = conf.low, ymax = conf.high, x = temp_centered + mean(temperature)),
              fill = "red", alpha = 0.3) +
  geom_line(data = pred_sd_temp,
            aes(x = temp_centered + mean(temperature), y = estimate, color = "Estimated")) +
  scale_color_manual(values = c("Simulated" = "darkgreen", "Estimated" = "red"), name = "Standard deviation") +
  labs(x = "Temperature", y = "Standard deviation")
```

`type = "disp"` retrieves predictions/slopes for the dispersion sub-model, exactly as `type = "conditional"` does for the mean.

## Simulating and fitting interactions on the standard deviation

```{r}
#| label: simulate-sd-interaction
log_sd_rain_slope <- 0.02
log_sd_interaction_effect <- 0.005

flower_measurements_3 <- tibble(
  temp_centered = temperature - mean(temperature),
  rain_centered = rain - mean(rain),
  petal_length = rnorm(
    n = n_flowers,
    mean = avg_petal_length +
      temperature_effect * temp_centered +
      rain_effect * rain_centered,
    sd = exp(log_sd_intercept +
      log_sd_temp_slope * temp_centered +
      log_sd_rain_slope * rain_centered +
      log_sd_interaction_effect * temp_centered * rain_centered))
)

mod_flower_measurements_3 <- glmmTMB(petal_length ~ temp_centered + rain_centered,
                                     dispformula = ~ temp_centered + rain_centered + temp_centered * rain_centered,
                                     data = flower_measurements_3)
```

We can interpret the interactive effects of rain and temperature on the standard deviation of petal length in the same way that we would for the average (see the Numerical x Numerical Interactions tutorial for worked examples). Another way to communicate the results: since the two parameters of the Normal distribution act together to generate the observations, we can predict new distributions of observations for different predictor values.

## Predicting new observations

```{r}
#| label: predicting-new-observations
#| fig-cap: "Predicted distributions of petal length at five reference rain and temperature levels."
predicted_observations <- predictions(mod_flower_measurements_3,
            newdata = datagrid(
              temp_centered = fivenum,
              rain_centered = fivenum
            ),
            type = "conditional") |>
  as_tibble() |>
  select(temp_centered, rain_centered, estimate) |>
  rename(mu = estimate) |>
  left_join(predictions(mod_flower_measurements_3,
            newdata = datagrid(
              temp_centered = fivenum,
              rain_centered = fivenum
            ),
            type = "disp") |>
  as_tibble() |>
  select(temp_centered, rain_centered, estimate) |>
  rename(sd = estimate)) |>
  rowwise() |>
  mutate(temp_centered = round(temp_centered),
         rain_centered = round(rain_centered),
         pred_obs = dist_normal(mu = mu, sd = sd))

# Build facet labels dynamically from the actual rounded temp_centered values
# (hardcoding them would silently show NA facet labels if the simulated data changes)
temp_labels <- setNames(
  paste0("temp = ", sort(unique(predicted_observations$temp_centered))),
  sort(unique(predicted_observations$temp_centered))
)

ggplot(predicted_observations, aes(xdist = pred_obs, color = as.factor(rain_centered))) +
  stat_slab(fill = NA, normalize = "xy") +
  scale_color_brewer(palette = "Accent") +
  facet_wrap(~ as.factor(temp_centered), scales = "free", labeller = as_labeller(temp_labels)) +
  labs(x = "Predicted petal length", color = "Rain (centered)") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
```

## Part 2 takeaways

1.  With two numerical predictors, the interaction coefficient is the *rate of change of one slope per unit of the other* — not a separate slope for each group
2.  Centering predictors makes main-effect coefficients interpretable as "effect at average conditions"
3.  Plot slopes across the moderator's range, not just the coefficient table
4.  Five-number-summary prediction plots are usually more honest than response surfaces (they retain uncertainty)
5.  Interactions can act on variance, not just the mean, via distributional regression (`dispformula`)

# Part 3 — Link Functions Create Implicit Interactions

So far, every interaction we've seen was written explicitly into the model formula (`X1 * X2`). Non-linear **link functions** (logit, log, ...) introduce a *second*, implicit source of interaction: even a model with **no interaction term at all** can show predictors whose response-scale effect depends on the values of other predictors. This happens because the link function bends a straight line (on the *link* scale) into a curve (on the *response* scale) — and how much a fixed push on the link scale moves the response depends on where you are on that curve.

## Case study: species occurrence

Simulated presence/absence of a species across 200 plots, as a function of tree cover, rain, and temperature — **no interaction terms**, fit with a binomial GLM (logit link).

```{r}
#| label: simulate-occurrence
n_plots <- 200
tree_cover <- dist_mixture(dist_beta(4, 0.8), dist_beta(0.8, 4), weights = c(0.5, 0.5))
avg_yearly_rain <- dist_gamma(shape = 3, rate = 0.7)
avg_yearly_temp <- dist_normal(mu = 20, sd = 10)

occurence_intercept <- 0.047
occurence_tree_effect <- 0.45
occurence_rain_effect <- 0.01
occurence_temp_effect <- 0.01

occ_data <- tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |>
  unnest(everything()) |>
  mutate(occurence = rbinom(n_plots, size = 1,
    prob = occurence_intercept +
      occurence_tree_effect * tree_cov +
      occurence_rain_effect * rain +
      occurence_temp_effect * temp))
```

## Fit the model

```{r}
#| label: fit-occurrence-model
mod_1 <- glmmTMB(occurence ~ tree_cov + rain + temp, family = binomial(link = "logit"), data = occ_data)
summary(mod_1)$coefficients$cond
```

We simulated the effect of tree cover to be **0.45** on the probability scale, but the model reports **\~1.87**. That's not an error — it's reporting the coefficient on the **log-odds** scale, not the probability scale.

## The logit link bends a line into an S-shape

```{r}
#| label: logistic-curve
#| fig-cap: "The logit link bends a straight line into an S-shape."
tibble(log_odds = seq(-6, 6, length.out = 400)) |>
  mutate(probability = plogis(log_odds)) |>
  ggplot(aes(x = log_odds, y = probability)) +
  geom_line() +
  labs(x = "log-odds scale (what the coefficient describes)",
       y = "probability scale (what we care about)")
```

Near the middle (p ≈ 0.5) the curve is steep: a small log-odds push moves probability a lot. Near the ends (p ≈ 0 or 1) the curve flattens: the same push barely moves probability. No manual derivative needed — `marginaleffects` converts automatically to the scale we actually care about.

## `marginaleffects` does this for us

```{r}
#| label: slopes-probability-scale
slopes(mod_1, variables = "tree_cov",
       newdata = datagrid(temp = mean(occ_data$temp), rain = mean(occ_data$rain)))
```

`slopes()` in this case asks: "at these covariate values, how much does *predicted probability* change for a small increase in tree cover?" — a probability-scale answer, directly comparable to how we simulated the data.

## The implicit interaction, visualized

```{r}
#| label: implicit-interaction-plot
#| fig-cap: "The tree-cover slope on the probability scale changes across rain x temp, though no interaction term exists in the model."
slopes(mod_1, variables = "tree_cov", newdata = datagrid(temp = fivenum, rain = fivenum)) |>
  ggplot() +
  geom_pointrange(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = 1)) +
  geom_vline(xintercept = occurence_tree_effect, alpha = .8) +
  facet_grid(rain ~ temp) +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank())
```

Because the amount of "bending" depends on where you are on the S-curve, tree cover's probability-scale effect depends on rain and temp too — even though we never told the model to include that interaction. Whether an interaction "exists" can depend on which scale you look at (additive/probability vs. multiplicative/log-odds) — see Spake et al. (2023) for more on this "Type-D" error.

## Response vs. link scale, across each moderator

```{r}
#| label: response-vs-link-scale
#| fig-cap: "Slope of tree_cov across temp (top) and rain (bottom), on the probability (left) vs. log-odds (right) scale."
slopes_by_temp_response <- slopes(mod_1,
  newdata = datagrid(temp = seq(min(occ_data$temp), max(occ_data$temp), length.out = 30),
                      rain = mean(occ_data$rain)),
  variables = "tree_cov", by = "temp", type = "response") |> as_tibble()

slopes_by_temp_link <- slopes(mod_1,
  newdata = datagrid(temp = seq(min(occ_data$temp), max(occ_data$temp), length.out = 30),
                      rain = mean(occ_data$rain)),
  variables = "tree_cov", by = "temp", type = "link") |> as_tibble()

slopes_by_rain_response <- slopes(mod_1,
  newdata = datagrid(rain = seq(min(occ_data$rain), max(occ_data$rain), length.out = 30),
                      temp = mean(occ_data$temp)),
  variables = "tree_cov", by = "rain", type = "response") |> as_tibble()

slopes_by_rain_link <- slopes(mod_1,
  newdata = datagrid(rain = seq(min(occ_data$rain), max(occ_data$rain), length.out = 30),
                      temp = mean(occ_data$temp)),
  variables = "tree_cov", by = "rain", type = "link") |> as_tibble()

plot_slope <- function(data, x, y_lab, scale_title) {
  ggplot(data) +
    geom_hline(yintercept = 0, color = "blue", linetype = 2) +
    geom_ribbon(aes(x = .data[[x]], ymin = conf.low, ymax = conf.high), alpha = 0.4, fill = "lightgrey") +
    geom_line(aes(x = .data[[x]], y = estimate)) +
    labs(y = y_lab, x = x, title = scale_title)
}

(plot_slope(slopes_by_temp_response, "temp", "Slope of tree_cov\n(probability scale)", "Response scale") +
 plot_slope(slopes_by_temp_link, "temp", "Slope of tree_cov\n(log-odds scale)", "Link scale")) /
(plot_slope(slopes_by_rain_response, "rain", "Slope of tree_cov\n(probability scale)", "Response scale") +
 plot_slope(slopes_by_rain_link, "rain", "Slope of tree_cov\n(log-odds scale)", "Link scale"))
```

**Link (log-odds) scale**: the slope of `tree_cov` is flat (\~1.87) across both moderators — exactly as it should be, since `mod_1`'s formula has no interaction term at all. **Response (probability) scale**: the slope visibly curves with temp (and more weakly with rain) — the implicit interaction made concrete, purely from the logit link's curvature. With three predictors, a focal variable's slope can no longer be shown across "the" moderator in one 2D plot — each panel above holds the *other* remaining predictor fixed at its mean. Alternatives: facet by the third variable, use small multiples at representative (e.g. five-number-summary) values, or report a single `avg_slopes()` that integrates over the observed distribution instead of fixing it.

## The "overall" effect of tree cover

```{r}
#| label: overall-tree-cover-effect
#| fig-cap: "Predicted probability of occurrence across tree_cov, averaged over the observed rain and temp values (counterfactual averaging)."
avg_predictions(mod_1, variables = "tree_cov", vcov = TRUE) |>
  ggplot() +
  geom_ribbon(aes(x = tree_cov, ymin = conf.low, ymax = conf.high), fill = "lightgrey", alpha = 0.4) +
  geom_line(aes(x = tree_cov, y = estimate)) +
  labs(y = "Predicted occurence")
```

`avg_predictions()`/`avg_slopes()` give one number (or curve) per focal value, integrating over rain and temp as observed in the sample — a marginal, "average effect" alternative to picking a single moderator value to condition on.

## 🧪 Exercise 5

Using `mod_1`:

1.  Plot the predicted trend of occurence as a function of tree cover for the unique combinations of the five-number summary of rain and temperature. Hint: use the `predictions()` function.
2.  Plot the prediction for the average effect of temperature.

*(\~20 minutes)*

```{r}
#| label: exercise-5-task-1

```

```{r}
#| label: exercise-5-task-2

```

## Reviewer 2 wants a table

```{r}
#| label: reviewer-2-table
# Average slopes on the probability scale for all three predictors
slopes_tbl_mod1 <- avg_slopes(mod_1, variables = c("tree_cov", "rain", "temp"), type = "response") |>
  as_tibble() |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  mutate(
    across(c(estimate, conf.low, conf.high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) |>
  mutate(CI_95 = glue::glue("[{conf.low}, {conf.high}]")) |>
  select(Predictor = term, Slope = estimate, `95% CI` = CI_95, `p-value` = p.value)

# Tjur's R2 as an extra row in the same 4-column layout
r2_row <- tibble(
  Predictor = "Tjur's R\u00b2",
  Slope = as.character(round(r2_tjur(mod_1), 3)),
  `95% CI` = "",
  `p-value` = ""
)

mod1_tbl <- bind_rows(
  slopes_tbl_mod1 |> mutate(across(everything(), as.character)),
  r2_row
)

mod1_tbl |>
  tt(caption = "Average slopes of occurrence on the probability scale, and Tjur's R2 for mod_1") |>
  style_tt(i = nrow(slopes_tbl_mod1) + 1, bold = TRUE, line = "t", line_color = "black", line_width = 0.1)
```

## Case study: species abundance

The same phenomenon shows up with **count** data. Abundance (when present) as a function of the same three predictors, fit with a Poisson GLM (log link), again **no interaction terms**.

```{r}
#| label: simulate-abundance
abundance_intercept <- exp(1)
abundance_tree_effect <- 4.5
abundance_rain_effect <- 0.1
abundance_temp_effect <- 0.1

abund_data <- tibble(
  tree_cov = generate(tree_cover, n_plots),
  rain = generate(avg_yearly_rain, n_plots),
  temp = generate(avg_yearly_temp, n_plots)
) |>
  unnest(everything()) |>
  mutate(abundance = rpois(n_plots, lambda =
    abundance_intercept +
    abundance_tree_effect * tree_cov +
    abundance_rain_effect * rain +
    abundance_temp_effect * temp))

mod_2 <- glmmTMB(abundance ~ tree_cov + rain + temp, family = poisson(link = "log"), data = abund_data)
summary(mod_2)$coefficients$cond
```

The log-scale `tree_cov` coefficient is **\~0.73**, not 4.5 — the same phenomenon as the logit link, just a different curve.

## The log link bends a line into a curve

```{r}
#| label: exponential-curve
#| fig-cap: "The log link bends a straight line into an exponential curve."
tibble(log_count = seq(-2, 6, length.out = 400)) |>
  mutate(count = exp(log_count)) |>
  ggplot(aes(x = log_count, y = count)) +
  geom_line() +
  labs(x = "log scale (what the coefficient describes)", y = "count scale (what we care about)")
```

## Two ways to report a log-link coefficient

- **Rate ratio**: a *constant* multiplicative effect, valid everywhere — "each unit increase in tree cover multiplies expected abundance by this factor, regardless of rain or temp"
- **Local slope**: an *additive*, count-scale effect ("this many more individuals"), but only valid near the specific covariate values used

## Computing the rate ratio

```{r}
#| label: rate-ratio
exp(fixef(mod_2)$cond["tree_cov"])  # rate ratio: ~2.08, i.e. a ~108% increase per unit tree cover
```

Both describe the same underlying model — pick the one that matches your audience (percent change vs. a concrete headcount).

## 🧪 Exercise 6

Using `mod_2`:

1.  Compute the rate ratio and the local count-scale slope for `temp`, holding `rain` at its minimum (a "drought" scenario) and `tree_cov` at its mean.
2.  Repeat holding `tree_cov` at a low vs. a high value (e.g., the 10th and 90th percentiles) — does the rate ratio change? Does the local slope?
3.  Explain, in your own words, why one of these two summaries stays constant across `tree_cov` and the other doesn't.

*(\~7 minutes)*

```{r}
#| label: exercise-6-task-1

```

```{r}
#| label: exercise-6-task-2

```

## Case study: occurrence *and* abundance together

In reality, a species first has to be **present** before we can measure its abundance. A **hurdle model** combines two sub-models fit jointly:

- A **binomial** (logit-link) component: does a plot clear the "hurdle" of zero? (occurrence)
- A **truncated count** component: given it's non-zero, how many individuals? (abundance when present)

Two ways to interpret: each part **separately** (what drives occurrence vs. abundance), or **combined** on the population scale (expected abundance across *all* plots = P(occurrence) × E\[abundance \| present\]).

## A simple hurdle example

```{r}
#| label: simple-hurdle-simulation
# Simple hurdle simulation: abundance_pres = counts if present; occurence = presence;
# abundance_pop = observed count in every plot (zero wherever the species is absent)
simple_pop_data <- tibble(
  tree_cov = generate(tree_cover, n_plots)
) |>
  unnest(everything()) |>
  mutate(abundance_pres = rpois(n_plots, lambda =
                     abundance_intercept +
                     abundance_tree_effect * tree_cov
                     ),
         occurence = rbinom(n_plots, size = 1,
                     prob = occurence_intercept +
                            occurence_tree_effect * tree_cov),
         abundance_pop = abundance_pres * occurence)

# Hurdle model: main formula is the (zero-truncated Poisson) abundance-when-present part;
# zi = ~ tree_cov models the probability of a structural zero (absence) with a logit link
mod_hu_simple <- glmmTMB(abundance_pop ~ tree_cov,
                         zi = ~ tree_cov,
                         family = truncated_poisson,
                         data = simple_pop_data)
```

## Population vs. present-only predictions

```{r}
#| label: population-vs-present-only
#| fig-cap: "Top: predicted abundance, population-scale (response) vs. when present (conditional). Bottom: predicted occurrence probability (1 - zprob)."
# type = "response": population-scale expected abundance across ALL plots (P(present) x count)
pred_resp_simple <- predictions(mod_hu_simple, type = "response") |> mutate(type = "response")

# type = "zprob": predicted probability of a structural zero (i.e. NOT occurring)
pred_prob_simple <- predictions(mod_hu_simple, type = "zprob")

# type = "conditional": expected abundance given the species is present
pred_cond_simple <- predictions(mod_hu_simple, type = "conditional") |> mutate(type = "conditional")

pred_abun_simple <- bind_rows(pred_resp_simple, pred_cond_simple)

plot_pred_abun_simple <- ggplot(pred_abun_simple) +
  geom_ribbon(aes(x = tree_cov, ymin = conf.low, ymax = conf.high, fill = type), alpha = 0.2) +
  geom_line(aes(x = tree_cov, y = estimate, color = type)) +
  labs(x = "Tree cover", y = "Abundance", color = "Prediction type", fill = "Prediction type")

# zprob is P(structural zero), so 1 - zprob is the occurrence probability
plot_pred_prob_simple <- ggplot(pred_prob_simple) +
  geom_ribbon(aes(x = tree_cov, ymin = 1 - conf.high, ymax = 1 - conf.low), fill = "lightgrey", alpha = 0.4) +
  geom_line(aes(x = tree_cov, y = 1 - estimate)) +
  labs(x = "Tree cover", y = "Occurence probability")

plot_pred_abun_simple / plot_pred_prob_simple
```

Predicted abundance when present (conditional) is always higher than population-scale abundance (response), since occurrence probability is below 100%. This is an implicit interaction **between the model's two parts**: tree cover's effect on population-scale abundance depends on its effect on both occurrence and abundance-when-present.

## Simulating with an explicit interaction

Prey and competitors interact in their effect on abundance (when present); tree cover drives occurrence only. Predictors are simulated through the model's own inverse link (`exp()`, `plogis()`) so they stay in a valid range.

```{r}
#| label: simulate-explicit-interaction
more_plots <- 4000
prey_abun <- dist_poisson(lambda = 6)
competitor_abun <- dist_poisson(lambda = 3)

pop_data <- tibble(
  tree_cov = generate(tree_cover, more_plots),
  prey = generate(prey_abun, more_plots),
  competitors = generate(competitor_abun, more_plots)
) |>
  unnest(everything()) |>
  mutate(
    abundance_pres = rpois(more_plots,
      lambda = exp(1 + 0.12 * prey - 0.15 * competitors - 0.045 * prey * competitors)),
    occurence = rbinom(more_plots, size = 1, prob = plogis(-2 + 0.45 * tree_cov)),
    abundance_pop = abundance_pres * occurence
  ) |>
  mutate(across(c(tree_cov, prey, competitors), ~ (.x - mean(.x)) / sd(.x), .names = "scl_{.col}"))
```

## Fitting the hurdle model

```{r}
#| label: fit-hurdle-model
mod_hu <- glmmTMB(
  abundance_pop ~ scl_prey * scl_competitors,
  zi = ~ scl_tree_cov,
  family = truncated_poisson,
  data = pop_data
)
```

`zi = ~ scl_tree_cov`: the zero-inflation (occurrence) part. The conditional formula's `scl_prey * scl_competitors` is an **explicit** interaction, layered on top of the log link's *implicit* one.

## The interaction on the count scale

```{r}
#| label: interaction-count-scale
#| fig-cap: "Count-scale effect of prey weakens (and reverses) as competitors increase."
slopes(mod_hu, variables = "scl_prey",
       newdata = datagrid(scl_competitors = seq(min(pop_data$scl_competitors),
                                                 max(pop_data$scl_competitors), length.out = 30)),
       type = "response") |>
  # unscale back to raw competitor counts for an interpretable x-axis
  mutate(competitors = scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors)) |>
  ggplot(aes(x = competitors, y = estimate)) +
  geom_hline(yintercept = 0, color = "blue", linetype = 2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.4, fill = "lightgrey") +
  geom_line() +
  labs(y = "Slope of scl_prey\n(count scale, individuals present)", x = "Competitors",
       title = "Count-scale effect of prey availability weakens as competitors increase")
```

## The interaction on the link scale

```{r}
#| label: interaction-link-scale
#| fig-cap: "Link-scale effect of prey is a straight, non-zero-crossing line in competitors — the explicit interaction, undistorted by the log link."
slopes(mod_hu, variables = "scl_prey",
       newdata = datagrid(scl_competitors = seq(min(pop_data$scl_competitors),
                                                 max(pop_data$scl_competitors), length.out = 30)),
       type = "link") |>
  # unscale back to raw competitor counts for an interpretable x-axis
  mutate(competitors = scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors)) |>
  ggplot(aes(x = competitors, y = estimate)) +
  geom_hline(yintercept = 0, color = "blue", linetype = 2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.4, fill = "lightgrey") +
  geom_line() +
  labs(y = "Slope of scl_prey\n(log scale)", x = "Competitors",
       title = "Link-scale effect of prey declines linearly with competitors")
```

Unlike the count-scale plot, this one is linear in competitors by construction — the model's formula (`scl_prey * scl_competitors`) is fit additively on the log scale, so the **explicit** interaction shows up here as a straight line, with no additional curvature from the log link itself.

## Isolating the interaction: rate ratios

To isolate something closer to "the interaction itself," we move to the rate-ratio scale: for each one-SD increase in competitors, by what *factor* does the effect of one extra SD of prey get multiplied? That's `exp()` of the interaction coefficient, and because it lives on the model's own additive (log) scale, it doesn't depend on where competitors are evaluated — unlike the count-scale slope.

```{r}
#| label: rate-ratio-interaction-table
rate_ratio_interaction_tbl <- avg_slopes(mod_hu, variables = "scl_prey",
                                          by = "scl_competitors",
                                          newdata = datagrid(scl_competitors = c(-1, 0, 1)),
                                          type = "link") |>
  as_tibble() |>
  mutate(rate_ratio = exp(estimate), rr_low = exp(conf.low), rr_high = exp(conf.high)) |>
  select(scl_competitors, rate_ratio, rr_low, rr_high, p.value) |>
  mutate(
    across(c(rate_ratio, rr_low, rr_high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) |>
  mutate(CI_95 = glue::glue("[{rr_low}, {rr_high}]")) |>
  select(`Competitors (scaled)` = scl_competitors, `Rate ratio for scl_prey` = rate_ratio, `95% CI` = CI_95, `p-value` = p.value)

rate_ratio_interaction_tbl |>
  tt(caption = "Rate ratio of scl_prey's effect on abundance, evaluated at three levels of scl_competitors")
```

Prey's rate ratio drops from \>1 (increasing abundance) at low competitor density to \<1 (decreasing abundance) at high competitor density.

## Predicted abundance, by competitor level

```{r}
#| label: predicted-abundance-by-competitor
#| fig-cap: "Abundance-when-present predictions across prey, one line per competitor five-number-summary level."
predictions(mod_hu,
  newdata = datagrid(
    scl_prey = seq(min(pop_data$scl_prey), max(pop_data$scl_prey), length.out = 30),
    scl_competitors = fivenum
  ),
  type = "conditional"
) |>
  mutate(prey = scl_prey * sd(pop_data$prey) + mean(pop_data$prey),
         competitors = round(scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors), 1)) |>
  ggplot() +
  geom_ribbon(aes(x = prey, ymin = conf.low, ymax = conf.high, fill = as.factor(competitors)), alpha = 0.2) +
  geom_line(aes(x = prey, y = estimate, color = as.factor(competitors))) +
  labs(y = "Predicted abundance\n(when present)", x = "Prey", color = "Competitors", fill = "Competitors")
```

At low competitor density, abundance climbs steeply with prey; by the time competitors are abundant, the prey curve is nearly flat — the same shrinking slope we saw before, now shown as predicted trends rather than slopes.

## Population-scale picture: low vs. high tree cover

```{r}
#| label: population-scale-by-tree-cover
#| fig-cap: "Population-scale (response) predictions: prey x competitors, faceted by low vs. high tree cover."
# Locate the two peaks of tree_cov's bimodal distribution, as scaled values
d <- density(pop_data$tree_cov)
peaks <- d$x[which(diff(sign(diff(d$y))) == -2) + 1]
low_tree_pop <- round(peaks[1], 2)
high_tree_pop <- round(peaks[length(peaks)], 2)
low_tree_pop_scl <- (low_tree_pop - mean(pop_data$tree_cov)) / sd(pop_data$tree_cov)
high_tree_pop_scl <- (high_tree_pop - mean(pop_data$tree_cov)) / sd(pop_data$tree_cov)

predictions(mod_hu,
  newdata = datagrid(
    scl_prey = seq(min(pop_data$scl_prey), max(pop_data$scl_prey), length.out = 30),
    scl_competitors = fivenum,
    scl_tree_cov = c(low_tree_pop_scl, high_tree_pop_scl)
  ),
  type = "response"
) |>
  mutate(prey = scl_prey * sd(pop_data$prey) + mean(pop_data$prey),
         competitors = round(scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors), 1),
         `Tree cover` = if_else(scl_tree_cov == low_tree_pop_scl, "Low tree cover", "High tree cover")) |>
  ggplot() +
  geom_ribbon(aes(x = prey, ymin = conf.low, ymax = conf.high, fill = as.factor(competitors)), alpha = 0.2) +
  geom_line(aes(x = prey, y = estimate, color = as.factor(competitors))) +
  facet_wrap(~ `Tree cover`) +
  labs(y = "Predicted population abundance", x = "Prey", color = "Competitors", fill = "Competitors")
```

Within each panel, the prey × competitors shape is the same. But the high-tree-cover panel sits above the low-tree-cover panel overall — more plots are occupied there. Tree cover mostly shifts *how many* plots have the species at all; prey and competitors jointly shape abundance within occupied plots.

## A symmetric interaction, viewed from the other predictor

```{r}
#| label: symmetric-interaction
#| fig-cap: "Same interaction, axes swapped: competitors on x, one line per prey five-number-summary level."
predictions(mod_hu,
  newdata = datagrid(
    scl_competitors = seq(min(pop_data$scl_competitors), max(pop_data$scl_competitors), length.out = 30),
    scl_prey = fivenum
  ),
  type = "conditional"
) |>
  mutate(competitors = scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors),
         prey = round(scl_prey * sd(pop_data$prey) + mean(pop_data$prey), 1)) |>
  ggplot() +
  geom_ribbon(aes(x = competitors, ymin = conf.low, ymax = conf.high, fill = as.factor(prey)), alpha = 0.2) +
  geom_line(aes(x = competitors, y = estimate, color = as.factor(prey))) +
  labs(y = "Predicted abundance\n(when present)", x = "Competitors", color = "Prey", fill = "Prey")
```

A two-way interaction is mathematically symmetric: `scl_prey:scl_competitors` is the same coefficient as `scl_competitors:scl_prey`. This view makes a different point than the prey-on-x-axis plot: regardless of prey availability, enough competitors drive abundance down to roughly the same low floor. Neither plot is "more correct" — reporting only one risks showing half the story (Spake et al. 2023).

## Average predictions: one curve per predictor

```{r}
#| label: average-predictions-one-curve
#| fig-cap: "Abundance-when-present, averaged over the full observed range of the other predictor."
avg_pred_prey_cond <- avg_predictions(mod_hu,
  newdata = datagrid(model = mod_hu,
                      scl_prey = seq(min(pop_data$scl_prey), max(pop_data$scl_prey), length.out = 30),
                      grid_type = "counterfactual"),
  by = "scl_prey", type = "conditional"
) |>
  mutate(prey = scl_prey * sd(pop_data$prey) + mean(pop_data$prey))

avg_pred_comp_cond <- avg_predictions(mod_hu,
  newdata = datagrid(model = mod_hu,
                      scl_competitors = seq(min(pop_data$scl_competitors), max(pop_data$scl_competitors), length.out = 30),
                      grid_type = "counterfactual"),
  by = "scl_competitors", type = "conditional"
) |>
  mutate(competitors = scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors))

p_avg_prey_cond <- avg_pred_prey_cond |>
  ggplot() +
  geom_ribbon(aes(x = prey, ymin = conf.low, ymax = conf.high), fill = "lightgrey", alpha = 0.4) +
  geom_line(aes(x = prey, y = estimate)) +
  labs(y = "Average predicted abundance\n(when present)", x = "Prey")

p_avg_comp_cond <- avg_pred_comp_cond |>
  ggplot() +
  geom_ribbon(aes(x = competitors, ymin = conf.low, ymax = conf.high), fill = "lightgrey", alpha = 0.4) +
  geom_line(aes(x = competitors, y = estimate)) +
  labs(y = "Average predicted abundance\n(when present)", x = "Competitors")

p_avg_prey_cond + p_avg_comp_cond
```

`datagrid(..., grid_type = "counterfactual")` pairs each focal value with every observed row, then averages — a true average over the sample rather than a slope evaluated at the other predictors' means. Averaged over all observed competitor density, prey still shows a positive effect; averaged over all observed prey, competitors show a diminishing-harm shape.

## Population-scale average, by tree cover

```{r}
#| label: population-scale-average-tree-cover
#| fig-cap: "Population-scale average curves, one pair per tree-cover peak."
avg_pred_prey_resp <- avg_predictions(mod_hu,
  newdata = datagrid(model = mod_hu,
                      scl_prey = seq(min(pop_data$scl_prey), max(pop_data$scl_prey), length.out = 30),
                      scl_tree_cov = c(low_tree_pop_scl, high_tree_pop_scl),
                      grid_type = "counterfactual"),
  by = c("scl_prey", "scl_tree_cov"), type = "response"
) |>
  mutate(prey = scl_prey * sd(pop_data$prey) + mean(pop_data$prey),
         `Tree cover` = if_else(scl_tree_cov == low_tree_pop_scl, "Low tree cover", "High tree cover"))

avg_pred_comp_resp <- avg_predictions(mod_hu,
  newdata = datagrid(model = mod_hu,
                      scl_competitors = seq(min(pop_data$scl_competitors), max(pop_data$scl_competitors), length.out = 30),
                      scl_tree_cov = c(low_tree_pop_scl, high_tree_pop_scl),
                      grid_type = "counterfactual"),
  by = c("scl_competitors", "scl_tree_cov"), type = "response"
) |>
  mutate(competitors = scl_competitors * sd(pop_data$competitors) + mean(pop_data$competitors),
         `Tree cover` = if_else(scl_tree_cov == low_tree_pop_scl, "Low tree cover", "High tree cover"))

p_avg_prey_resp <- avg_pred_prey_resp |>
  ggplot(aes(x = prey, y = estimate, color = `Tree cover`, fill = `Tree cover`)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line() +
  labs(y = "Average predicted\npopulation abundance", x = "Prey")

p_avg_comp_resp <- avg_pred_comp_resp |>
  ggplot(aes(x = competitors, y = estimate, color = `Tree cover`, fill = `Tree cover`)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  geom_line() +
  labs(y = "Average predicted\npopulation abundance", x = "Competitors")

p_avg_prey_resp + p_avg_comp_resp
```

The shapes barely change between tree-cover panels — the prey × competitors interaction lives entirely in the abundance-when-present part. What changes is the overall height: high-tree-cover populations sit consistently above low-tree-cover ones, a pure occupancy effect, cleanly separated from the abundance-when-present shape.

## A full rate-ratio table, by model component

```{r}
#| label: full-rate-ratio-table
# Conditional-model rate ratio for scl_prey, at three levels of scl_competitors
prey_rr_tbl <- avg_slopes(mod_hu, variables = "scl_prey", by = "scl_competitors",
                           newdata = datagrid(scl_competitors = c(-1, 0, 1)),
                           type = "link") |>
  mutate(rate_ratio = exp(estimate), rr_low = exp(conf.low), rr_high = exp(conf.high),
         Term = glue::glue("scl_prey (at scl_competitors = {scl_competitors} SD)")) |>
  select(Term, rate_ratio, rr_low, rr_high, p.value)

# Conditional-model rate ratio for scl_competitors, at three levels of scl_prey
comp_rr_tbl <- avg_slopes(mod_hu, variables = "scl_competitors", by = "scl_prey",
                           newdata = datagrid(scl_prey = c(-1, 0, 1)),
                           type = "link") |>
  mutate(rate_ratio = exp(estimate), rr_low = exp(conf.low), rr_high = exp(conf.high),
         Term = glue::glue("scl_competitors (at scl_prey = {scl_prey} SD)")) |>
  select(Term, rate_ratio, rr_low, rr_high, p.value)

# The interaction itself: ratio of rate ratios per 1-SD increase in the moderator
hu_confint <- confint(mod_hu)
hu_coef_cond <- summary(mod_hu)$coefficients$cond
hu_coef_zi <- summary(mod_hu)$coefficients$zi

interaction_rr_tbl <- tibble(
  Term = "scl_prey:scl_competitors (interaction)",
  rate_ratio = exp(hu_coef_cond["scl_prey:scl_competitors", "Estimate"]),
  rr_low = exp(hu_confint["cond.scl_prey:scl_competitors", "2.5 %"]),
  rr_high = exp(hu_confint["cond.scl_prey:scl_competitors", "97.5 %"]),
  p.value = hu_coef_cond["scl_prey:scl_competitors", "Pr(>|z|)"]
)

# Zero-inflation-model odds ratio for scl_tree_cov (no moderator in this model)
zi_rr_tbl <- tibble(
  Term = "scl_tree_cov (zero-inflation)",
  rate_ratio = exp(hu_coef_zi["scl_tree_cov", "Estimate"]),
  rr_low = exp(hu_confint["zi.scl_tree_cov", "2.5 %"]),
  rr_high = exp(hu_confint["zi.scl_tree_cov", "97.5 %"]),
  p.value = hu_coef_zi["scl_tree_cov", "Pr(>|z|)"]
)

hu_rr_tbl <- bind_rows(
  prey_rr_tbl |> mutate(Component = "Conditional (abundance, when present)"),
  comp_rr_tbl |> mutate(Component = "Conditional (abundance, when present)"),
  interaction_rr_tbl |> mutate(Component = "Conditional (abundance, when present)"),
  zi_rr_tbl |> mutate(Component = "Zero-inflation (occupancy, as an odds ratio)")
) |>
  relocate(Component) |>
  mutate(
    across(c(rate_ratio, rr_low, rr_high), ~ round(.x, 3)),
    p.value = if_else(p.value < 0.001, "<0.001", as.character(round(p.value, 3))),
    CI_95 = glue::glue("[{rr_low}, {rr_high}]")
  ) |>
  select(Component, Term, `Rate/odds ratio` = rate_ratio, `95% CI` = CI_95, `p-value` = p.value)

hu_rr_tbl |>
  tt(caption = "Standardized rate ratios (conditional model) and odds ratio (zero-inflation model) for mod_hu") |>
  style_tt(i = which(hu_rr_tbl$Component != dplyr::lag(hu_rr_tbl$Component, default = "")) - 1,
           j = 1:5, line = "b", line_color = "black", line_width = 0.05)
```

Prey's rate ratio crosses 1 somewhere around average competitor density (why its raw `summary()` estimate looked like noise); competitors suppress abundance more as prey increases; the interaction row (\~0.81) is the single number behind both patterns, and it's the *same* number whichever predictor you treat as the moderator — the symmetry point from a few sections back, visible directly in the table. Tree cover's odds ratio in the zero-inflation part tells a simpler, interaction-free story about occupancy alone.

## 🧪 Exercise 7 / Homework

Now apply what we learned today to your own data!

```{r}
#| label: exercise-7

```

## Part 3 takeaways

1.  Non-linear link functions (logit, log) create an **implicit interaction**: every predictor's response-scale effect depends on the values of all other predictors, even with no interaction term in the formula
2.  Raw GLM coefficients live on the link scale; `marginaleffects::slopes()` converts them to the natural (probability, count) scale for you
3.  Multiplicative summaries (rate ratios, odds ratios) stay constant across the range of other predictors; additive, response-scale summaries do not
4.  Whether an interaction "exists" can depend on the scale you're measuring on (Spake et al. 2023) — this applies just as much to *implicit* link-function interactions as to explicit interaction terms
5.  **Hurdle/zero-inflated models** separate "does it occur" from "how much, given it occurs" — interpret each part separately, or combine them (`type = "response"`) for a population-scale answer
6.  Interactions are mathematically **symmetric**: report (or at least check) both "A moderated by B" and "B moderated by A"

## Further reading

- Full write-ups with additional detail and 3D visualizations: [Part 1](../../posts/interactions_categorical_numerical/index.qmd), [Part 2](../../posts/interactions_numerical_numerical/index.qmd), [Part 3](../../posts/interactions_link_functions/index.qmd)
- `marginaleffects` documentation: <https://marginaleffects.com/>
- Spake et al. (2023), on interactions in GLMs: <https://onlinelibrary.wiley.com/doi/full/10.1111/brv.12939>
- Mize (2019), best practices for interaction effects: <https://sociologicalscience.com/articles-v6-4-81/>